---
title: "Complete Clinical Analysis Walkthrough"
subtitle: "From Kaplan-Meier baseline to validated multivariable model"
author: "John Ehrlinger"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Complete Clinical Analysis Walkthrough}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = has_pkg
)
```

This vignette demonstrates the **complete analytical workflow** for a
temporal parametric hazard analysis, mirroring the disciplined sequence used
in the original SAS HAZARD system:

1. **Nonparametric baseline** --- Kaplan-Meier life table
2. **Shape fitting** --- start simple, build to multiphase
3. **Variable screening** --- calibrate covariates, assess functional form
4. **Multivariable model** --- covariate entry with fixed shapes
5. **Prediction** --- patient-specific risk profiles
6. **Validation** --- decile-of-risk calibration

This corresponds to the SAS programs `ac.*` → `hz.*` → `lg.*` → `hm.*` →
`hp.*` → `hs.*`.

We use the **AVC** dataset (310 patients, atrioventricular canal repair)
which has rich covariates and two identifiable hazard phases.

## Data preparation

```{r}
#| label: setup
library(TemporalHazard)
if (has_ggplot2) library(ggplot2)

data(avc)
avc <- na.omit(avc)
str(avc)
```

The AVC dataset contains `r if (exists("avc", inherits = TRUE)) nrow(avc) else 310` patients after removing incomplete
cases.  Key covariates include age at repair (`age`, in months), NYHA
functional class (`status`), presence of malalignment (`mal`), and
interventricular communication (`com_iv`).

## Step 1: Nonparametric baseline

Before fitting any parametric model, establish the empirical survival curve
using the Kaplan-Meier estimator.  This is the benchmark against which all
parametric fits will be compared.

`hzr_kaplan()` wraps `survival::survfit()` and adds logit-transformed exact
confidence limits that respect the `[0, 1]` boundary --- more accurate in
the tails than the default Greenwood intervals.  This matches the SAS
`kaplan.sas` macro output structure.

```{r}
#| label: km-life-table
km <- hzr_kaplan(time = avc$int_dead, status = avc$dead)
head(km)
```

The returned data frame is the life table: one row per event time with
`n_risk`, `n_event`, `survival`, logit-transformed 95% CLs
(`cl_lower` / `cl_upper`), and the Nelson-Aalen cumulative hazard.
Plotting just requires the survival column:

```{r}
#| label: fig-km-baseline
#| fig-cap: "Kaplan-Meier survival estimate for AVC patients (n = 310)"
#| eval: !expr has_pkg && has_ggplot2
km_df <- data.frame(
  time     = km$time,
  survival = km$survival * 100,
  lower    = km$cl_lower * 100,
  upper    = km$cl_upper * 100
)

ggplot(km_df, aes(time, survival)) +
  geom_step(colour = "#D55E00", linewidth = 0.8) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              stat = "identity", alpha = 0.15, fill = "#D55E00") +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

The KM curve shows a sharp early drop (operative mortality) followed by a
roughly constant attrition rate.  This two-phase pattern suggests an early
CDF phase plus a constant phase --- no obvious late rising hazard.

## Step 2: Shape fitting --- simple to complex

### 2a. Single-phase Weibull

Start with the simplest parametric model to establish a baseline fit.

```{r}
#| label: weibull-fit
fit_weib <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data  = avc,
  dist  = "weibull",
  theta = c(mu = 0.05, nu = 0.5),
  fit   = TRUE
)
summary(fit_weib)
```

The Weibull forces a monotone hazard shape.  Let's overlay it on the
Kaplan-Meier to see where it fits well and where it doesn't.

```{r}
#| label: fig-weibull-overlay
#| fig-cap: "Single Weibull vs. Kaplan-Meier"
#| eval: !expr has_pkg && has_ggplot2
t_grid <- seq(0.01, max(avc$int_dead) * 0.9, length.out = 200)
surv_weib <- predict(fit_weib,
                     newdata = data.frame(time = t_grid),
                     type = "survival") * 100

ggplot() +
  geom_step(data = km_df, aes(time, survival),
            colour = "#D55E00", linewidth = 0.6) +
  geom_line(data = data.frame(time = t_grid, survival = surv_weib),
            aes(time, survival), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

The single Weibull typically misses the sharp early drop --- it compromises
between the early and late time frames.

### 2b. Two-phase model (early CDF + constant)

Based on the KM shape, fit an early phase (resolving operative risk) plus
a constant phase (background attrition).

```{r}
#| label: multiphase-fit
fit_mp <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)
summary(fit_mp)
```

Note the use of `fixed = "shapes"` --- we fix the temporal shape parameters
and only estimate the scale (log_mu) for each phase.  This matches the
standard HAZARD workflow: shapes are set from clinical knowledge or
preliminary exploration, then scales and covariates are estimated.

```{r}
#| label: fig-multiphase-overlay
#| fig-cap: "Two-phase parametric model vs. Kaplan-Meier"
#| eval: !expr has_pkg && has_ggplot2
surv_mp <- predict(fit_mp,
                   newdata = data.frame(time = t_grid),
                   type = "survival") * 100

ggplot() +
  geom_step(data = km_df, aes(time, survival),
            colour = "#D55E00", linewidth = 0.6) +
  geom_line(data = data.frame(time = t_grid, survival = surv_mp),
            aes(time, survival), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

### 2c. Decomposed hazard

Visualize the per-phase contributions to the cumulative hazard.

```{r}
#| label: fig-decomposed
#| fig-cap: "Phase decomposition of cumulative hazard"
#| eval: !expr has_pkg && has_ggplot2
decomp <- predict(fit_mp,
                  newdata = data.frame(time = t_grid),
                  type = "cumulative_hazard",
                  decompose = TRUE)

decomp_df <- data.frame(
  time  = t_grid,
  total = decomp[, "total"],
  early = decomp[, "early"],
  const = decomp[, "constant"]
)

ggplot(decomp_df, aes(x = time)) +
  geom_line(aes(y = total), linewidth = 0.9, colour = "black") +
  geom_line(aes(y = early), linewidth = 0.7, colour = "#E69F00",
            linetype = "dashed") +
  geom_line(aes(y = const), linewidth = 0.7, colour = "#56B4E9",
            linetype = "dashed") +
  labs(x = "Months after AVC repair", y = "Cumulative hazard") +
  theme_minimal()
```

The early phase contributes most of its risk in the first few months, then
levels off.  The constant phase accumulates linearly over time.

## Step 3: Variable screening

Before entering covariates into the hazard model, screen them for
association with the outcome.  This step corresponds to the `lg.*` programs
in the SAS workflow.

### 3a. Univariable logistic screening

Use simple logistic regression of the event indicator on each covariate to
get a quick ranking by strength of association.

```{r}
#| label: screening
covariates <- c("age", "status", "mal", "com_iv", "orifice",
                "inc_surg", "opmos")

screen_results <- data.frame(
  variable = covariates,
  coef     = numeric(length(covariates)),
  p_value  = numeric(length(covariates))
)

for (i in seq_along(covariates)) {
  v <- covariates[i]
  fml <- as.formula(paste("dead ~", v))
  lg <- glm(fml, data = avc, family = binomial)
  s <- summary(lg)$coefficients
  if (nrow(s) >= 2) {
    screen_results$coef[i] <- s[2, 1]
    screen_results$p_value[i] <- s[2, 4]
  }
}

screen_results <- screen_results[order(screen_results$p_value), ]
screen_results$significant <- ifelse(screen_results$p_value < 0.05,
                                      "*", "")
screen_results
```

### 3b. Functional form assessment

For continuous covariates that appear significant, examine whether the
relationship with outcome is monotone on the logit scale.  `hzr_calibrate()`
implements the SAS `logit.sas` / `logitgr.sas` macros: bin the continuous
covariate into quantile groups, compute the observed event rate per bin,
and transform it to the logit scale.

```{r}
#| label: age-calibrate
cal_age <- hzr_calibrate(x = avc$age, event = avc$dead,
                          groups = 10, link = "logit")
cal_age
```

Plot the logit-transformed event rate against the bin centre (`mean`
column).  A straight line means the covariate can enter the model
untransformed; monotone-but-curved shapes suggest a log or square-root
transform; U-shapes call for a quadratic term.

```{r}
#| label: fig-calibrate-age
#| fig-cap: "Decile calibration: age vs. logit(P(death))"
#| eval: !expr has_pkg && has_ggplot2
ggplot(cal_age, aes(mean, link_value)) +
  geom_point(size = 3, colour = "#0072B2") +
  geom_line(colour = "#0072B2", alpha = 0.4) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "#D55E00", linetype = "dashed") +
  labs(x = "Age at repair (months, decile midpoint)",
       y = "logit(P(death))") +
  theme_minimal()
```

The blue points are the observed decile logits; the dashed red line is a
linear reference.  Deviations from linearity flag where a transform is
warranted.

## Step 4: Multivariable model

### 4a. Manual specification

Enter the significant covariates from screening into the two-phase
hazard model directly.

```{r}
#| label: multivariable-fit
fit_mv <- hazard(
  survival::Surv(int_dead, dead) ~ age + status + mal + com_iv,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)
summary(fit_mv)
```

The coefficient table shows phase-specific covariate effects.  A positive
coefficient means higher hazard (worse prognosis).  Interpretation:
covariates scale the phase-specific hazard multiplicatively via
exp(x * beta).

### 4b. Automated stepwise selection

The manual approach works when screening has already narrowed the
candidate pool, but with a larger pool it helps to let the model
choose.  `hzr_stepwise()` runs forward, backward, or two-way
selection against an existing `hazard` fit, scoring candidates with
Wald p-values or ΔAIC.  Defaults match SAS `PROC HAZARD`
(`SLENTRY = 0.30`, `SLSTAY = 0.20`).

We start from a baseline with no covariates, offer the same screened
candidate pool, and let the algorithm decide.  For multiphase models
the scope is a named list of one-sided formulas keyed by phase, so
each phase can advertise its own candidate set; stepping operates
per `(variable, phase)` pair, letting a covariate enter one phase
and not another.  Single-distribution models accept either a flat
one-sided formula or a character vector of names.

```{r}
#| label: stepwise-forward
#| eval: !expr has_pkg
base_mp <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 3, maxit = 500)
)

fit_step <- hzr_stepwise(
  base_mp,
  scope     = list(
    early    = ~ age + status + mal + com_iv,
    constant = ~ age + status + mal + com_iv
  ),
  data      = avc,
  direction = "both",
  criterion = "wald",
  trace     = FALSE,
  control   = list(n_starts = 2, maxit = 500)
)
fit_step
```

The `$steps` table records every accepted action with its Wald
statistic and p-value:

```{r}
#| eval: !expr has_pkg
fit_step$steps[, c("step_num", "action", "variable", "phase",
                   "p_value", "aic")]
```

The final model is the fit at the top of the object, reachable
through the usual hazard accessors:

```{r}
#| eval: !expr has_pkg
logLik_manual <- fit_mv$fit$objective
logLik_step   <- fit_step$fit$objective
c(manual = logLik_manual, stepwise = logLik_step,
  aic_manual = 2 * length(fit_mv$fit$theta) - 2 * logLik_manual,
  aic_step   = 2 * length(fit_step$fit$theta) - 2 * logLik_step)
```

When the screening and stepwise agree on the same covariate set the
log-likelihoods match exactly; when stepwise selects a leaner model
AIC drops.  For an AIC-driven run use `criterion = "aic"`; for a
forward-only sweep `direction = "forward"`.  See `?hzr_stepwise` for
the full argument surface including `force_in`, `force_out`, and the
`max_move` oscillation guard.

## Step 5: Prediction

### 5a. Baseline survival curve

Generate the survival curve for a reference patient (median age, NYHA
class 1, no malalignment, no interventricular communication).

`predict(..., se.fit = TRUE)` adds a delta-method standard error and 95%
confidence band around every prediction (log(-log) scale for survival so
the band is guaranteed to lie in `[0, 1]`).  This is the parametric
analogue of the KM confidence limits from Step 1 --- a patient-specific
uncertainty estimate the nonparametric curve cannot provide.

```{r}
#| label: fig-prediction
#| fig-cap: "Reference patient survival with 95% delta-method CI and KM overlay"
#| eval: !expr has_pkg && has_ggplot2
ref_patient <- data.frame(
  time   = t_grid,
  age    = median(avc$age),
  status = 1,
  mal    = 0,
  com_iv = 0
)

surv_ref <- predict(fit_mv, newdata = ref_patient,
                    type = "survival", se.fit = TRUE)

ref_df <- data.frame(
  time     = t_grid,
  survival = surv_ref$fit * 100,
  lower    = surv_ref$lower * 100,
  upper    = surv_ref$upper * 100
)

ggplot() +
  geom_step(data = km_df, aes(time, survival),
            colour = "#D55E00", linewidth = 0.6, alpha = 0.7) +
  geom_ribbon(data = ref_df, aes(time, ymin = lower, ymax = upper),
              fill = "#0072B2", alpha = 0.15) +
  geom_line(data = ref_df, aes(time, survival),
            colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

### 5b. Sensitivity analysis --- risk factor comparison

Compare survival curves for a low-risk vs. high-risk patient profile to
visualize the effect of covariates on long-term outcome.

```{r}
#| label: fig-sensitivity
#| fig-cap: "Survival by risk profile: low-risk (blue) vs. high-risk (red)"
#| eval: !expr has_pkg && has_ggplot2
low_risk <- data.frame(
  time   = t_grid,
  age    = quantile(avc$age, 0.25),
  status = 1,
  mal    = 0,
  com_iv = 0
)

high_risk <- data.frame(
  time   = t_grid,
  age    = quantile(avc$age, 0.90),
  status = 4,
  mal    = 1,
  com_iv = 1
)

surv_lo <- predict(fit_mv, newdata = low_risk,
                   type = "survival", se.fit = TRUE)
surv_hi <- predict(fit_mv, newdata = high_risk,
                   type = "survival", se.fit = TRUE)

sens_df <- data.frame(
  time     = rep(t_grid, 2),
  survival = c(surv_lo$fit, surv_hi$fit) * 100,
  lower    = c(surv_lo$lower, surv_hi$lower) * 100,
  upper    = c(surv_lo$upper, surv_hi$upper) * 100,
  profile  = rep(c("Low risk", "High risk"), each = length(t_grid))
)

ggplot(sens_df, aes(time, survival, colour = profile, fill = profile)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Low risk" = "#0072B2",
                                  "High risk" = "#D55E00")) +
  scale_fill_manual(values = c("Low risk" = "#0072B2",
                                "High risk" = "#D55E00")) +
  labs(x = "Months after AVC repair", y = "Survival (%)",
       colour = NULL, fill = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Non-overlapping confidence bands between the two profiles indicate the
risk-factor combination is well-identified; overlap around the bands'
centre-of-mass flags where the model can't distinguish the profiles with
the available sample size.

## Step 6: Validation --- decile-of-risk calibration

Partition patients into deciles by predicted risk and compare observed
vs. expected event rates.  Good calibration means the two track each
other.  This implements the workflow of the SAS `deciles.hazard.sas`
macro.

```{r}
#| label: fig-deciles
#| fig-cap: "Observed vs. expected event rates by risk decile"
#| eval: !expr has_pkg && has_ggplot2
cal <- hzr_deciles(fit_mv, time = max(avc$int_dead))
print(cal)
```

```{r}
#| label: fig-decile-plot
#| fig-cap: "Decile calibration: observed (bars) vs. expected (points) event rates"
#| eval: !expr has_pkg && has_ggplot2
ggplot(cal, aes(x = group)) +
  geom_col(aes(y = observed_rate), fill = "#56B4E9", alpha = 0.7) +
  geom_point(aes(y = expected_rate), colour = "#D55E00", size = 3) +
  geom_line(aes(y = expected_rate), colour = "#D55E00", linewidth = 0.5) +
  scale_x_continuous(breaks = seq_len(nrow(cal))) +
  labs(x = "Risk decile (1 = lowest)", y = "Event rate",
       caption = paste("Overall chi-sq =",
                        round(attr(cal, "overall")$chi_sq, 2),
                        ", p =",
                        format.pval(attr(cal, "overall")$p_value,
                                     digits = 3))) +
  theme_minimal()
```

A non-significant overall chi-square statistic (p > 0.05) indicates
adequate calibration --- the model's predictions are consistent with
observed event rates across the risk spectrum.

### Conservation of events check

The conservation-of-events principle states that a well-fitting model
should predict the same total number of events as actually observed.
`hzr_gof()` tracks this cumulatively over time --- the residual
(expected minus observed) should stay near zero.

```{r}
#| label: gof-summary
#| eval: !expr has_pkg
gof <- hzr_gof(fit_mv)
print(gof)
```

```{r}
#| label: fig-gof-survival
#| fig-cap: "Parametric (blue) vs. Kaplan-Meier (orange) survival"
#| eval: !expr has_pkg && has_ggplot2
ggplot(gof, aes(x = time)) +
  geom_step(aes(y = km_surv * 100), colour = "#D55E00", linewidth = 0.6) +
  geom_line(aes(y = par_surv * 100), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

```{r}
#| label: fig-gof-events
#| fig-cap: "Cumulative observed (orange) vs. expected (blue) events"
#| eval: !expr has_pkg && has_ggplot2
ggplot(gof, aes(x = time)) +
  geom_line(aes(y = cum_observed), colour = "#D55E00", linewidth = 0.8) +
  geom_line(aes(y = cum_expected), colour = "#0072B2",
            linewidth = 0.8, linetype = "dashed") +
  labs(x = "Months after AVC repair", y = "Cumulative events") +
  theme_minimal()
```

```{r}
#| label: fig-gof-residual
#| fig-cap: "Conservation-of-events residual (expected minus observed) over time"
#| eval: !expr has_pkg && has_ggplot2
ggplot(gof, aes(x = time, y = residual)) +
  geom_line(linewidth = 0.7, colour = "grey30") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(x = "Months after AVC repair",
       y = "Residual (expected - observed)") +
  theme_minimal()
```

A residual that hovers near zero across the full time range indicates the
model conserves events well.  Persistent positive residual means the model
overpredicts risk; persistent negative means it underpredicts.

## Why not Cox regression?

The temporal parametric approach offers several advantages over Cox
proportional hazards for this type of analysis:

| Feature | Cox (`coxph`) | TemporalHazard |
|:---|:---:|:---:|
| Proportional hazards required | Yes | No |
| Baseline hazard specified | No | Yes (parametric) |
| Multiple hazard phases | No | Yes (additive) |
| Patient-specific survival prediction | Approximate | Exact |
| Smooth extrapolation beyond data | No | Yes |
| Conservation of events check | No | Yes (`hzr_gof()`) |
| Interval censoring | Not standard | Supported |

The key insight is that many clinical outcomes exhibit **non-proportional
hazards**: the risk factors that dominate early mortality (e.g., surgical
complexity) differ from those driving late attrition (e.g., age, comorbidity).
The multiphase model captures this naturally through phase-specific
covariate effects.

## Summary

The complete analytical workflow follows a disciplined sequence:

1. **Kaplan-Meier baseline** --- establish the empirical survival pattern
2. **Shape fitting** --- match the temporal shape, simple to complex
3. **Variable screening** --- logistic screening, LOESS functional form
4. **Multivariable model** --- enter covariates with fixed shapes
5. **Prediction** --- reference curves, sensitivity analysis
6. **Validation** --- decile calibration, conservation-of-events check

This sequence ensures that the temporal shape is established before
covariates are introduced, and that the final model is validated against
the observed data.  See `vignette("getting-started")` for the minimal
API workflow and `vignette("mf-mathematical-foundations")` for the
mathematical details.
