---
title: "Complete Clinical Analysis Walkthrough"
subtitle: "From Kaplan-Meier baseline to validated multivariable model"
author: "John Ehrlinger"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Complete Clinical Analysis Walkthrough}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = has_pkg
)
```

This vignette demonstrates the **complete analytical workflow** for a
temporal parametric hazard analysis, mirroring the disciplined sequence used
in the original SAS HAZARD system:

1. **Nonparametric baseline** --- Kaplan-Meier life table
2. **Shape fitting** --- start simple, build to multiphase
3. **Variable screening** --- calibrate covariates, assess functional form
4. **Multivariable model** --- covariate entry with fixed shapes
5. **Prediction** --- patient-specific risk profiles
6. **Validation** --- decile-of-risk calibration

This corresponds to the SAS programs `ac.*` → `hz.*` → `lg.*` → `hm.*` →
`hp.*` → `hs.*`.

We use the **AVC** dataset (310 patients, atrioventricular canal repair)
which has rich covariates and two identifiable hazard phases.

## Data preparation

```{r}
#| label: setup
library(TemporalHazard)
if (has_ggplot2) library(ggplot2)

data(avc)
avc <- na.omit(avc)
str(avc)
```

The AVC dataset contains `r nrow(avc)` patients after removing incomplete
cases.  Key covariates include age at repair (`age`, in months), NYHA
functional class (`status`), presence of malalignment (`mal`), and
interventricular communication (`com_iv`).

## Step 1: Nonparametric baseline

Before fitting any parametric model, establish the empirical survival curve
using the Kaplan-Meier estimator.  This is the benchmark against which all
parametric fits will be compared.

```{r}
#| label: fig-km-baseline
#| fig-cap: "Kaplan-Meier survival estimate for AVC patients (n = 310)"
#| eval: !expr has_pkg && has_ggplot2
km <- survival::survfit(survival::Surv(int_dead, dead) ~ 1, data = avc)

km_df <- data.frame(
  time     = km$time,
  survival = km$surv * 100,
  lower    = km$lower * 100,
  upper    = km$upper * 100
)

ggplot(km_df, aes(time, survival)) +
  geom_step(colour = "#D55E00", linewidth = 0.8) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              stat = "identity", alpha = 0.15, fill = "#D55E00") +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

The KM curve shows a sharp early drop (operative mortality) followed by a
roughly constant attrition rate.  This two-phase pattern suggests an early
CDF phase plus a constant phase --- no obvious late rising hazard.

## Step 2: Shape fitting --- simple to complex

### 2a. Single-phase Weibull

Start with the simplest parametric model to establish a baseline fit.

```{r}
#| label: weibull-fit
fit_weib <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data  = avc,
  dist  = "weibull",
  theta = c(mu = 0.05, nu = 0.5),
  fit   = TRUE
)
summary(fit_weib)
```

The Weibull forces a monotone hazard shape.  Let's overlay it on the
Kaplan-Meier to see where it fits well and where it doesn't.

```{r}
#| label: fig-weibull-overlay
#| fig-cap: "Single Weibull vs. Kaplan-Meier"
#| eval: !expr has_pkg && has_ggplot2
t_grid <- seq(0.01, max(avc$int_dead) * 0.9, length.out = 200)
surv_weib <- predict(fit_weib,
                     newdata = data.frame(time = t_grid),
                     type = "survival") * 100

ggplot() +
  geom_step(data = km_df, aes(time, survival),
            colour = "#D55E00", linewidth = 0.6) +
  geom_line(data = data.frame(time = t_grid, survival = surv_weib),
            aes(time, survival), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

The single Weibull typically misses the sharp early drop --- it compromises
between the early and late time frames.

### 2b. Two-phase model (early CDF + constant)

Based on the KM shape, fit an early phase (resolving operative risk) plus
a constant phase (background attrition).

```{r}
#| label: multiphase-fit
fit_mp <- hazard(
  survival::Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)
summary(fit_mp)
```

Note the use of `fixed = "shapes"` --- we fix the temporal shape parameters
and only estimate the scale (log_mu) for each phase.  This matches the
standard HAZARD workflow: shapes are set from clinical knowledge or
preliminary exploration, then scales and covariates are estimated.

```{r}
#| label: fig-multiphase-overlay
#| fig-cap: "Two-phase parametric model vs. Kaplan-Meier"
#| eval: !expr has_pkg && has_ggplot2
surv_mp <- predict(fit_mp,
                   newdata = data.frame(time = t_grid),
                   type = "survival") * 100

ggplot() +
  geom_step(data = km_df, aes(time, survival),
            colour = "#D55E00", linewidth = 0.6) +
  geom_line(data = data.frame(time = t_grid, survival = surv_mp),
            aes(time, survival), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

### 2c. Decomposed hazard

Visualize the per-phase contributions to the cumulative hazard.

```{r}
#| label: fig-decomposed
#| fig-cap: "Phase decomposition of cumulative hazard"
#| eval: !expr has_pkg && has_ggplot2
decomp <- predict(fit_mp,
                  newdata = data.frame(time = t_grid),
                  type = "cumulative_hazard",
                  decompose = TRUE)

decomp_df <- data.frame(
  time  = t_grid,
  total = decomp[, "total"],
  early = decomp[, "early"],
  const = decomp[, "constant"]
)

ggplot(decomp_df, aes(x = time)) +
  geom_line(aes(y = total), linewidth = 0.9, colour = "black") +
  geom_line(aes(y = early), linewidth = 0.7, colour = "#E69F00",
            linetype = "dashed") +
  geom_line(aes(y = const), linewidth = 0.7, colour = "#56B4E9",
            linetype = "dashed") +
  labs(x = "Months after AVC repair", y = "Cumulative hazard") +
  theme_minimal()
```

The early phase contributes most of its risk in the first few months, then
levels off.  The constant phase accumulates linearly over time.

## Step 3: Variable screening

Before entering covariates into the hazard model, screen them for
association with the outcome.  This step corresponds to the `lg.*` programs
in the SAS workflow.

### 3a. Univariable logistic screening

Use simple logistic regression of the event indicator on each covariate to
get a quick ranking by strength of association.

```{r}
#| label: screening
covariates <- c("age", "status", "mal", "com_iv", "orifice",
                "inc_surg", "opmos")

screen_results <- data.frame(
  variable = covariates,
  coef     = numeric(length(covariates)),
  p_value  = numeric(length(covariates))
)

for (i in seq_along(covariates)) {
  v <- covariates[i]
  fml <- as.formula(paste("dead ~", v))
  lg <- glm(fml, data = avc, family = binomial)
  s <- summary(lg)$coefficients
  if (nrow(s) >= 2) {
    screen_results$coef[i] <- s[2, 1]
    screen_results$p_value[i] <- s[2, 4]
  }
}

screen_results <- screen_results[order(screen_results$p_value), ]
screen_results$significant <- ifelse(screen_results$p_value < 0.05,
                                      "*", "")
screen_results
```

### 3b. Functional form assessment

For continuous covariates that appear significant, examine whether the
relationship with outcome is linear on the logit scale using LOESS
smoothing.  Non-linear patterns suggest a transformation may be needed.

```{r}
#| label: fig-loess-age
#| fig-cap: "LOESS smooth: age vs. mortality probability"
#| eval: !expr has_pkg && has_ggplot2
ggplot(avc, aes(age, dead)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
              colour = "#0072B2", fill = "#0072B2", alpha = 0.2) +
  labs(x = "Age at repair (months)", y = "P(death)") +
  theme_minimal()
```

If the LOESS curve is roughly linear (or monotone), the covariate can enter
the model untransformed.  S-shaped or U-shaped curves suggest
transformations (log, quadratic) may improve the fit.

## Step 4: Multivariable model

Enter the significant covariates into the two-phase hazard model.  With
the current package, variable selection is done manually --- identify the
promising covariates from screening, then fit the full model.

```{r}
#| label: multivariable-fit
fit_mv <- hazard(
  survival::Surv(int_dead, dead) ~ age + status + mal + com_iv,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)
summary(fit_mv)
```

The coefficient table shows phase-specific covariate effects.  A positive
coefficient means higher hazard (worse prognosis).  Interpretation:
covariates scale the phase-specific hazard multiplicatively via
exp(x * beta).

## Step 5: Prediction

### 5a. Baseline survival curve

Generate the survival curve for a reference patient (median age, NYHA
class 1, no malalignment, no interventricular communication).

```{r}
#| label: fig-prediction
#| fig-cap: "Reference patient survival with Kaplan-Meier overlay"
#| eval: !expr has_pkg && has_ggplot2
ref_patient <- data.frame(
  time   = t_grid,
  age    = median(avc$age),
  status = 1,
  mal    = 0,
  com_iv = 0
)

surv_ref <- predict(fit_mv, newdata = ref_patient,
                    type = "survival") * 100

ggplot() +
  geom_step(data = km_df, aes(time, survival),
            colour = "#D55E00", linewidth = 0.6, alpha = 0.7) +
  geom_line(data = data.frame(time = t_grid, survival = surv_ref),
            aes(time, survival), colour = "#0072B2", linewidth = 0.8) +
  labs(x = "Months after AVC repair", y = "Survival (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal()
```

### 5b. Sensitivity analysis --- risk factor comparison

Compare survival curves for a low-risk vs. high-risk patient profile to
visualize the effect of covariates on long-term outcome.

```{r}
#| label: fig-sensitivity
#| fig-cap: "Survival by risk profile: low-risk (blue) vs. high-risk (red)"
#| eval: !expr has_pkg && has_ggplot2
low_risk <- data.frame(
  time   = t_grid,
  age    = quantile(avc$age, 0.25),
  status = 1,
  mal    = 0,
  com_iv = 0
)

high_risk <- data.frame(
  time   = t_grid,
  age    = quantile(avc$age, 0.90),
  status = 4,
  mal    = 1,
  com_iv = 1
)

surv_lo <- predict(fit_mv, newdata = low_risk, type = "survival") * 100
surv_hi <- predict(fit_mv, newdata = high_risk, type = "survival") * 100

sens_df <- data.frame(
  time = rep(t_grid, 2),
  survival = c(surv_lo, surv_hi),
  profile = rep(c("Low risk", "High risk"), each = length(t_grid))
)

ggplot(sens_df, aes(time, survival, colour = profile)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Low risk" = "#0072B2",
                                  "High risk" = "#D55E00")) +
  labs(x = "Months after AVC repair", y = "Survival (%)",
       colour = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Step 6: Validation --- decile-of-risk calibration

Partition patients into deciles by predicted risk and compare observed
vs. expected event rates.  Good calibration means the two track each
other.  This implements the workflow of the SAS `deciles.hazard.sas`
macro.

```{r}
#| label: fig-deciles
#| fig-cap: "Observed vs. expected event rates by risk decile"
#| eval: !expr has_pkg && has_ggplot2
cal <- hzr_deciles(fit_mv, time = max(avc$int_dead))
print(cal)
```

```{r}
#| label: fig-decile-plot
#| fig-cap: "Decile calibration: observed (bars) vs. expected (points) event rates"
#| eval: !expr has_pkg && has_ggplot2
ggplot(cal, aes(x = group)) +
  geom_col(aes(y = observed_rate), fill = "#56B4E9", alpha = 0.7) +
  geom_point(aes(y = expected_rate), colour = "#D55E00", size = 3) +
  geom_line(aes(y = expected_rate), colour = "#D55E00", linewidth = 0.5) +
  scale_x_continuous(breaks = seq_len(nrow(cal))) +
  labs(x = "Risk decile (1 = lowest)", y = "Event rate",
       caption = paste("Overall chi-sq =",
                        round(attr(cal, "overall")$chi_sq, 2),
                        ", p =",
                        format.pval(attr(cal, "overall")$p_value,
                                     digits = 3))) +
  theme_minimal()
```

A non-significant overall chi-square statistic (p > 0.05) indicates
adequate calibration --- the model's predictions are consistent with
observed event rates across the risk spectrum.

## Why not Cox regression?

The temporal parametric approach offers several advantages over Cox
proportional hazards for this type of analysis:

| Feature | Cox (`coxph`) | TemporalHazard |
|:---|:---:|:---:|
| Proportional hazards required | Yes | No |
| Baseline hazard specified | No | Yes (parametric) |
| Multiple hazard phases | No | Yes (additive) |
| Patient-specific survival prediction | Approximate | Exact |
| Smooth extrapolation beyond data | No | Yes |
| Conservation of events check | No | Planned |
| Interval censoring | Not standard | Supported |

The key insight is that many clinical outcomes exhibit **non-proportional
hazards**: the risk factors that dominate early mortality (e.g., surgical
complexity) differ from those driving late attrition (e.g., age, comorbidity).
The multiphase model captures this naturally through phase-specific
covariate effects.

## Summary

The complete analytical workflow follows a disciplined sequence:

1. **Kaplan-Meier baseline** --- establish the empirical survival pattern
2. **Shape fitting** --- match the temporal shape, simple to complex
3. **Variable screening** --- logistic screening, LOESS functional form
4. **Multivariable model** --- enter covariates with fixed shapes
5. **Prediction** --- reference curves, sensitivity analysis
6. **Validation** --- decile calibration, parametric vs. KM overlay

This sequence ensures that the temporal shape is established before
covariates are introduced, and that the final model is validated against
the observed data.  See `vignette("getting-started")` for the minimal
API workflow and `vignette("mf-mathematical-foundations")` for the
mathematical details.
