---
title: "Example: Hazard Sensitivity & Scenarios (hs.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Hazard Sensitivity & Scenarios (hs.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `hs.*` SAS files: sensitivity analyses and
scenario predictions from a fitted hazard model, typically comparing outcomes
across different patient profiles or covariate settings.

---

## hs.death.AVC.hm1 — Sensitivity analysis, AVC model 1

**SAS source:** `examples/hs.death.AVC.hm1.sas`

**Purpose:** Evaluate sensitivity of the AVC death model to covariate changes.
Compares predicted survival at median vs extreme covariate values.

### R translation (stub — M2 predict types and M3 scenario helpers required)

```{r hs-death-avc-hm1}
# fit_hm1 loaded from hm.death.AVC example

# Define reference and scenario covariate profiles
ref_profile <- data.frame(
  AGE = median(avcs$AGE), COM_IV = 0, MAL = 0,
  OPMOS = median(avcs$OPMOS), OP_AGE = median(avcs$OP_AGE),
  STATUS = 2, INC_SURG = 0, ORIFICE = 0
)

# Scenario: older patient, higher NYHA, important anomaly
scen_profile <- ref_profile
scen_profile$AGE    <- quantile(avcs$AGE, 0.9)
scen_profile$STATUS <- 4
scen_profile$MAL    <- 1

# eta_ref  <- predict(fit_hm1, newdata = data.matrix(ref_profile),
#                     type = "linear_predictor")
# eta_scen <- predict(fit_hm1, newdata = data.matrix(scen_profile),
#                     type = "linear_predictor")
```

---

## hs.dthar.TGA.setup / hs.dthar.TGA.hs1 — Arrhythmia after TGA repair

**SAS sources:** `examples/hs.dthar.TGA.setup.sas`, `examples/hs.dthar.TGA.hs1.sas`

**Purpose:** Two-step analysis for arrhythmia after transposition of the great
arteries (TGA) repair. The `setup` file prepares the dataset and fits the
hazard model; `hs1` runs the scenario predictions.

### R translation (stub)

```{r hs-dthar-tga-setup}
# Step 1: Fit model (from hs.dthar.TGA.setup.sas)
# tga <- read.table("data/tga", ...)

# fit_tga <- hazard(
#   time    = tga$time_ar,
#   status  = tga$arrhythmia,
#   x       = data.matrix(tga[, c("age", "repair_type", "bypass_time")]),
#   dist    = "weibull"
# )
```

```{r hs-dthar-tga-hs1}
# Step 2: Scenario predictions (from hs.dthar.TGA.hs1.sas)
# scenarios <- expand.grid(
#   age         = c(1, 5, 10),
#   repair_type = c(0, 1),
#   bypass_time = median(tga$bypass_time)
# )

# eta_scenarios <- predict(fit_tga,
#                          newdata = data.matrix(scenarios),
#                          type = "linear_predictor")
# hv_hazard(fit_tga, groups = scenarios)  # hvtiPlotR
```
