---
title: "Example: Hazard Prediction (hp.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Hazard Prediction (hp.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(hvtiRhazard)
library(hvtiPlotR)
```

These examples correspond to the `hp.*` SAS files: prediction from a fitted
hazard model including survival curves, patient-level risk, and group
comparisons. These map to `predict.hazard()` and `hvtiPlotR` plotting output.

---

## hp.death.AVC — Predicted survival after AVC repair

**SAS source:** `examples/hp.death.AVC.sas`

**Purpose:** Compute and plot predicted survival and hazard curves from the
fitted AVC model.

### SAS (original)

```sas
%HAZPRED(
PROC HAZARD DATA=OUTEST IN=AVCS OUT=HPOUT;
     TIME INT_DEAD;
     PREDICT SURVIVAL HAZARD;
);
```

### R translation (stub — M2 predict types required)

```{r hp-death-avc}
# fit_hm loaded from hm.death.AVC example

# Survival and hazard predictions on a time grid
time_grid <- seq(0, max(avcs$INT_DEAD, na.rm = TRUE), length.out = 500)

# hz_pred  <- predict(fit_hm, newdata = ..., type = "hazard")
# surv_pred <- predict(fit_hm, newdata = ..., type = "survival")  # M2+

# Plot with hvtiPlotR
# hv_hazard(fit_hm, time_max = max(time_grid))
```

---

## hp.death.AVC.hm1 / hp.death.AVC.hm2 — Predictions from two models

**SAS sources:** `examples/hp.death.AVC.hm1.sas`, `examples/hp.death.AVC.hm2.sas`

**Purpose:** Compare predicted hazard curves from two fitted multivariable
models (e.g., alternative covariate specifications).

### R translation (stub)

```{r hp-avc-hm1-hm2}
# fit_hm1, fit_hm2 — two separately fitted hazard() objects

# hv_hazard_compare(list(Model1 = fit_hm1, Model2 = fit_hm2))  # M3+
```

---

## hp.death.COMPARISON — Model comparison

**SAS source:** `examples/hp.death.COMPARISON.sas`

**Purpose:** Overlay predicted survival/hazard from multiple models or covariate
scenarios on a single plot.

### R translation (stub)

```{r hp-death-comparison}
# predict() on each fitted model, then bind and plot
# Requires: predict(fit, type = "survival")   (M2+)
```

---

## hp.dthip.PAIVS.time — Prediction with time-varying risk

**SAS source:** `examples/hp.dthip.PAIVS.time.sas`

**Purpose:** Predict patient-specific hazard over time for the PAIVS (Pulmonary
Atresia with Intact Ventricular Septum) dataset.

### R translation (stub)

```{r hp-dthip-paivs}
# paivs <- read.table("data/paivs", ...)

# fit_paivs <- hazard(time = paivs$time, status = paivs$dead, ...)
# eta <- predict(fit_paivs, type = "linear_predictor")
```
