---
title: "Example: Bootstrap Confidence Intervals (bs.*)"
format: html
---

<!--
%\VignetteIndexEntry{Example: Bootstrap Confidence Intervals (bs.*)}
%\VignetteEngine{knitr::quarto}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| label: setup
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `bs.*` SAS files: bootstrap confidence
intervals for hazard function estimates and model summaries. Bootstrap CI is
the primary uncertainty quantification method used in the HAZARD workflow.

---

## bs.death.AVC — Bootstrap CI for AVC death hazard

**SAS source:** `examples/bs.death.AVC.sas`

**Purpose:** Generate bootstrap confidence intervals for the parametric hazard
function fitted to AVC death data. Each bootstrap replicate refits the model
and accumulates prediction curves; the CI is then summarized across replicates.

### SAS (original)

```sas
%BOOTSTRAP(
  DATA    = AVCS,
  HAZEST  = EXAMPLES.HMDEATH,
  N       = 200,
  SEED    = 12345,
  OUT     = BSOUT
);
```

### R translation (runnable M3 example)

```{r}
#| label: bs-death-avc
set.seed(3201)
n <- 280
avcs <- data.frame(
  INT_DEAD = rexp(n, rate = 0.6),
  AGE = rnorm(n, 6, 3),
  COM_IV = rbinom(n, 1, 0.2),
  MAL = rbinom(n, 1, 0.2),
  OPMOS = runif(n, 0, 18),
  OP_AGE = runif(n, 0, 12),
  STATUS = sample(1:4, n, replace = TRUE),
  INC_SURG = rbinom(n, 1, 0.12),
  ORIFICE = rbinom(n, 1, 0.1)
)

lp <- -2.5 + 0.05 * avcs$AGE + 0.6 * avcs$COM_IV + 0.4 * avcs$MAL +
  0.25 * (avcs$STATUS >= 3) + 0.4 * avcs$INC_SURG
avcs$DEAD <- rbinom(n, 1, plogis(lp))

X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                          "STATUS", "INC_SURG", "ORIFICE")])

fit_hm <- hazard(
  time = avcs$INT_DEAD,
  status = avcs$DEAD,
  x = X,
  theta = c(mu = 0.25, nu = 1.10, rep(0, ncol(X))),
  dist = "weibull",
  fit = TRUE,
  control = list(maxit = 400)
)

time_grid <- data.frame(time = seq(0.1, 4.0, length.out = 40))

# Bootstrap loop (smaller n for vignette runtime).
set.seed(12345)
n_boot <- 60

boot_curves <- replicate(n_boot, simplify = FALSE, {
  idx    <- sample(nrow(avcs), replace = TRUE)
  boot_d <- avcs[idx, ]
  boot_x <- data.matrix(boot_d[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                                   "STATUS", "INC_SURG", "ORIFICE")])

  out <- tryCatch({
    boot_fit <- hazard(
      time   = boot_d$INT_DEAD,
      status = boot_d$DEAD,
      x      = boot_x,
      theta  = coef(fit_hm),
      dist   = "weibull",
      fit    = TRUE,
      control = list(maxit = 250)
    )

    predict(boot_fit, newdata = cbind(time_grid, as.data.frame(boot_x[1:nrow(time_grid), ])),
            type = "survival")
  }, error = function(e) {
    rep(NA_real_, nrow(time_grid))
  })

  out
})

length(boot_curves)
```

---

## bs.death.AVC.summary — Bootstrap summary statistics

**SAS source:** `examples/bs.death.AVC.summary.sas`

**Purpose:** Compute percentile-based confidence bands from bootstrap replicates
and produce publication-ready hazard function plots with shaded CI.

### R translation (runnable M3 example)

```{r}
#| label: bs-avc-summary
ci_bands <- do.call(rbind, boot_curves)
ci_bands <- ci_bands[rowSums(is.na(ci_bands)) == 0, , drop = FALSE]
median_surv <- apply(ci_bands, 2, median)
lower <- apply(ci_bands, 2, quantile, probs = 0.025)
upper <- apply(ci_bands, 2, quantile, probs = 0.975)

bootstrap_summary <- data.frame(
  time = time_grid$time,
  surv_median = median_surv,
  surv_lower = lower,
  surv_upper = upper
)

head(bootstrap_summary)
```
