---
title: "Example: Logistic/Gompertz Exploratory Plots (lg.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Logistic/Gompertz Exploratory Plots (lg.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(hvtiRhazard)
library(hvtiPlotR)
```

These examples correspond to the `lg.*` SAS files: exploratory logistic and
Gompertz plots used to identify covariate transformations and preliminary
functional forms *before* fitting the parametric hazard model. These are
diagnostic/exploratory tools, not the primary analysis.

---

## lg.death.AVC — Logistic/Gompertz plots, AVC death

**SAS source:** `examples/lg.death.AVC.sas`

**Purpose:** For each candidate covariate, fit a logistic (early hazard) or
Gompertz (late hazard) model and plot the estimated relationship. This guides
covariate transformation decisions (log, power, inverse) before the hazard
model.

### SAS (original)

```sas
%LOGIT(
  DATA  = AVCS,
  EVENT = DEAD,
  VARS  = AGE STATUS INC_SURG OPMOS MAL COM_IV ORIFICE OP_AGE
);
```

### R translation (stub — M3 exploratory helpers required)

```{r lg-death-avc}
# avcs <- read.table("data/avc", ...)

candidates <- c("AGE", "STATUS", "INC_SURG", "OPMOS",
                "MAL", "COM_IV", "ORIFICE", "OP_AGE")

# Fit logistic model for each candidate covariate
logit_fits <- lapply(candidates, function(var) {
  fmla <- as.formula(paste("DEAD ~", var))
  glm(fmla, data = avcs, family = binomial())
})
names(logit_fits) <- candidates

# Inspect: which covariates show monotone relationships?
lapply(logit_fits, function(f) coef(summary(f)))

# Plot covariate effects (hvtiPlotR exploratory helper — M3)
# hv_logit_explorer(avcs, event = "DEAD", vars = candidates)
```

---

## lg.dead30d.AVC — 30-day mortality logistic model

**SAS source:** `examples/lg.dead30d.AVC.sas`

**Purpose:** Logistic regression for early (30-day) mortality after AVC repair.
Identifies covariates driving operative risk.

### R translation (stub)

```{r lg-dead30d-avc}
# Define 30-day event indicator
# avcs$dead30 <- as.integer(avcs$DEAD == 1 & avcs$INT_DEAD <= 1)  # 1 month

# fit_30d <- glm(dead30 ~ AGE + STATUS + COM_IV + MAL,
#               data = avcs, family = binomial())
# summary(fit_30d)
```

---

## lg.deadlate.AVC — Late mortality logistic model

**SAS source:** `examples/lg.deadlate.AVC.sas`

**Purpose:** Logistic regression restricted to patients surviving 30 days.
Identifies covariates driving late (constant-phase) hazard.

### R translation (stub)

```{r lg-deadlate-avc}
# Restrict to survivors of operative period
# avcs_late <- avcs[avcs$INT_DEAD > 1 | avcs$DEAD == 0, ]
# avcs_late$dead_late <- as.integer(avcs_late$DEAD == 1)

# fit_late <- glm(dead_late ~ INC_SURG + ORIFICE + STATUS,
#                data = avcs_late, family = binomial())
# summary(fit_late)
```

### Note on lg.* → hz.* workflow

The typical HAZARD analysis sequence is:

1. `lg.*` — identify covariate transformations and functional forms
2. `ac.*` — nonparametric estimate for baseline characterization
3. `hz.*` — fit parametric hazard function shape
4. `hm.*` — stepwise multivariable model
5. `hp.*` — predictions from final model
6. `bs.*` — bootstrap confidence intervals
7. `hs.*` — sensitivity and scenario analyses
