---
title: "Example: Bootstrap Confidence Intervals (bs.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Bootstrap Confidence Intervals (bs.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `bs.*` SAS files: bootstrap confidence
intervals for hazard function estimates and model summaries. Bootstrap CI is
the primary uncertainty quantification method used in the HAZARD workflow.

---

## bs.death.AVC — Bootstrap CI for AVC death hazard

**SAS source:** `examples/bs.death.AVC.sas`

**Purpose:** Generate bootstrap confidence intervals for the parametric hazard
function fitted to AVC death data. Each bootstrap replicate refits the model
and accumulates prediction curves; the CI is then summarized across replicates.

### SAS (original)

```sas
%BOOTSTRAP(
  DATA    = AVCS,
  HAZEST  = EXAMPLES.HMDEATH,
  N       = 200,
  SEED    = 12345,
  OUT     = BSOUT
);
```

### R translation (stub — M3 bootstrap harness required)

```{r bs-death-avc}
# fit_hm loaded from hm.death.AVC example

# Bootstrap loop (200 replicates, seed for reproducibility)
set.seed(12345)
n_boot <- 200

boot_curves <- replicate(n_boot, simplify = FALSE, {
  idx    <- sample(nrow(avcs), replace = TRUE)
  boot_d <- avcs[idx, ]

  boot_fit <- hazard(
    time   = boot_d$INT_DEAD,
    status = boot_d$DEAD,
    x      = data.matrix(boot_d[, c("AGE", "COM_IV", "MAL", "OPMOS",
                                    "OP_AGE", "STATUS", "INC_SURG", "ORIFICE")]),
    theta  = coef(fit_hm),     # warm start from fitted model (M2+)
    dist   = "weibull",
    control = list(condition = 14)
  )

  # predict(boot_fit, type = "survival")  # M2+ survival predictions
})

# Summarize CI across replicates and plot
# hv_hazard(fit_hm, boot_curves = boot_curves)  # hvtiPlotR
```

---

## bs.death.AVC.summary — Bootstrap summary statistics

**SAS source:** `examples/bs.death.AVC.summary.sas`

**Purpose:** Compute percentile-based confidence bands from bootstrap replicates
and produce publication-ready hazard function plots with shaded CI.

### R translation (stub)

```{r bs-avc-summary}
# Summarize boot_curves into 2.5/97.5 percentile bands
# ci_bands <- do.call(rbind, lapply(boot_curves, function(x) x$survival))
# lower <- apply(ci_bands, 2, quantile, probs = 0.025)
# upper <- apply(ci_bands, 2, quantile, probs = 0.975)

# Plot with hvtiPlotR confidence bands
# hv_hazard(fit_hm,
#           ci_lower = lower,
#           ci_upper = upper)
```
