---
title: "Example: Hazard Sensitivity & Scenarios (hs.*)"
format: html
---

<!--
%\VignetteIndexEntry{Example: Hazard Sensitivity & Scenarios (hs.*)}
%\VignetteEngine{quarto::html}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| label: setup
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `hs.*` SAS files: sensitivity analyses and
scenario predictions from a fitted hazard model, typically comparing outcomes
across different patient profiles or covariate settings.

---

## hs.death.AVC.hm1 — Sensitivity analysis, AVC model 1

**SAS source:** `examples/hs.death.AVC.hm1.sas`

**Purpose:** Evaluate sensitivity of the AVC death model to covariate changes.
Compares predicted survival at median vs extreme covariate values.

### R translation (runnable M3 example)

```{r}
#| label: hs-death-avc-hm1
set.seed(4201)
n <- 300
avcs <- data.frame(
  INT_DEAD = rexp(n, rate = 0.65),
  AGE = rnorm(n, mean = 6, sd = 3),
  COM_IV = rbinom(n, 1, 0.2),
  MAL = rbinom(n, 1, 0.25),
  OPMOS = runif(n, 0, 18),
  OP_AGE = runif(n, 0, 12),
  STATUS = sample(1:4, n, replace = TRUE),
  INC_SURG = rbinom(n, 1, 0.12),
  ORIFICE = rbinom(n, 1, 0.1)
)

lp <- -2.6 + 0.05 * avcs$AGE + 0.65 * avcs$COM_IV + 0.35 * avcs$MAL +
  0.22 * (avcs$STATUS >= 3) + 0.45 * avcs$INC_SURG
avcs$DEAD <- rbinom(n, 1, plogis(lp))

X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                          "STATUS", "INC_SURG", "ORIFICE")])

fit_hm1 <- hazard(
  time = avcs$INT_DEAD,
  status = avcs$DEAD,
  x = X,
  theta = c(mu = 0.25, nu = 1.10, rep(0, ncol(X))),
  dist = "weibull",
  fit = TRUE,
  control = list(maxit = 400)
)

# Define reference and scenario covariate profiles
ref_profile <- data.frame(
  AGE = median(avcs$AGE), COM_IV = 0, MAL = 0,
  OPMOS = median(avcs$OPMOS), OP_AGE = median(avcs$OP_AGE),
  STATUS = 2, INC_SURG = 0, ORIFICE = 0
)

# Scenario: older patient, higher NYHA, important anomaly
scen_profile <- ref_profile
scen_profile$AGE    <- quantile(avcs$AGE, 0.9)
scen_profile$STATUS <- 4
scen_profile$MAL    <- 1

pred_ref <- cbind(data.frame(time = 1.5), ref_profile)
pred_scen <- cbind(data.frame(time = 1.5), scen_profile)

eta_ref  <- predict(fit_hm1, newdata = pred_ref, type = "linear_predictor")
eta_scen <- predict(fit_hm1, newdata = pred_scen, type = "linear_predictor")
s_ref <- predict(fit_hm1, newdata = pred_ref, type = "survival")
s_scen <- predict(fit_hm1, newdata = pred_scen, type = "survival")

data.frame(
  profile = c("reference", "scenario"),
  linear_predictor = c(eta_ref, eta_scen),
  survival_t1_5 = c(s_ref, s_scen)
)
```

---

## hs.dthar.TGA.setup / hs.dthar.TGA.hs1 — Arrhythmia after TGA repair

**SAS sources:** `examples/hs.dthar.TGA.setup.sas`, `examples/hs.dthar.TGA.hs1.sas`

**Purpose:** Two-step analysis for arrhythmia after transposition of the great
arteries (TGA) repair. The `setup` file prepares the dataset and fits the
hazard model; `hs1` runs the scenario predictions.

### R translation (runnable M3 example)

```{r}
#| label: hs-dthar-tga-setup
set.seed(4202)
n <- 260
tga <- data.frame(
  time_ar = rexp(n, rate = 0.55),
  age = runif(n, 0.5, 12),
  repair_type = rbinom(n, 1, 0.45),
  bypass_time = rnorm(n, mean = 110, sd = 25)
)
tga$arrhythmia <- rbinom(
  n,
  1,
  plogis(-3.0 + 0.06 * tga$age + 0.55 * tga$repair_type + 0.008 * tga$bypass_time)
)

tga_x <- data.matrix(tga[, c("age", "repair_type", "bypass_time")])

fit_tga <- hazard(
  time    = tga$time_ar,
  status  = tga$arrhythmia,
  x       = tga_x,
  theta   = c(log_lambda = log(0.18), rep(0, ncol(tga_x))),
  dist    = "exponential",
  fit     = TRUE,
  control = list(maxit = 350)
)

fit_tga
```

```{r}
#| label: hs-dthar-tga-hs1
scenarios <- expand.grid(
  age         = c(1, 5, 10),
  repair_type = c(0, 1),
  bypass_time = median(tga$bypass_time),
  time = c(1, 3, 5)
)

eta_scenarios <- predict(fit_tga, newdata = scenarios, type = "linear_predictor")
s_scenarios <- predict(fit_tga, newdata = scenarios, type = "survival")

cbind(scenarios, linear_predictor = eta_scenarios, survival = s_scenarios)
```
