---
title: "Example: Estimate Hazard Function (hz.*)"
format: html
---

<!--
%\VignetteIndexEntry{Example: Estimate Hazard Function (hz.*)}
%\VignetteEngine{knitr::quarto}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| label: setup
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `hz.*` SAS files: univariable hazard function
estimation with no covariates. The hazard function shape is fit to
time-to-event data to characterize early, late, and constant hazard phases.

---

## hz.death.AVC — Death after AVC repair

**SAS source:** `examples/hz.death.AVC.sas`

**Purpose:** Estimate the parametric hazard function for death after repair of
atrioventricular septal defects. No covariates — shape parameters only.

### SAS (original)

```sas
%HAZARD(
PROC HAZARD DATA=AVCS NOCOV NOCOR CONDITION=14;
     TIME INT_DEAD;
     EVENT DEAD;
     PARMS MUE=0.2 THALF=0.15 NU=1.4 M=1;
);
```

### R translation (runnable M3 example)

```{r}
#| label: hz-death-avc
# Self-contained synthetic cohort (replace with real AVCS data in production).
set.seed(1201)
n <- 220
avcs <- data.frame(
  INT_DEAD = rexp(n, rate = 0.7),
  DEAD = rbinom(n, size = 1, prob = 0.55)
)

fit_hz <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = NULL,
  theta   = c(mu = 0.30, nu = 1.15),
  dist    = "weibull",
  fit     = TRUE,
  control = list(maxit = 300)
)

fit_hz

# Predict survival curve on a follow-up grid.
hz_grid <- data.frame(time = seq(0.05, 4.0, length.out = 40))
hz_grid$survival <- predict(fit_hz, newdata = hz_grid, type = "survival")
head(hz_grid)
```

---

## hz.deadp.KUL — Death after KUL procedures

**SAS source:** `examples/hz.deadp.KUL.sas`

**Purpose:** Estimate hazard function for the KUL (Katholieke Universiteit
Leuven) cardiac surgery dataset.

### R translation (runnable M3 example)

```{r}
#| label: hz-deadp-kul
set.seed(1202)
n <- 260
kul <- data.frame(
  follow_time = rexp(n, rate = 0.5),
  dead = rbinom(n, size = 1, prob = 0.45)
)

fit_kul <- hazard(
  time    = kul$follow_time,
  status  = kul$dead,
  theta   = c(log_lambda = log(0.18)),
  dist    = "exponential",
  fit     = TRUE,
  control = list(maxit = 300)
)

fit_kul

kul_grid <- data.frame(time = seq(0.05, 5.0, length.out = 50))
kul_grid$cumhaz <- predict(fit_kul, newdata = kul_grid, type = "cumulative_hazard")
head(kul_grid)
```

---

## hz.te123.OMC / hz.tm123.OMC — Time-to-event with multiple phases (OMC)

**SAS sources:** `examples/hz.te123.OMC.sas`, `examples/hz.tm123.OMC.sas`

**Purpose:** Three-phase hazard model (early, constant, late) for the OMC
(Ontario Multicenter) dataset. `te123` uses event-time input; `tm123` uses
the same model with a time-magnitude parameterization.

### R translation (runnable M3 example)

```{r}
#| label: hz-te123-omc
set.seed(1203)
n <- 280
omc <- data.frame(
  lo = runif(n, min = 0.05, max = 2.0),
  width = runif(n, min = 0.05, max = 1.0)
)
omc$hi <- omc$lo + omc$width
omc$status <- sample(c(1, 0, -1, 2), size = n, replace = TRUE,
                     prob = c(0.40, 0.35, 0.10, 0.15))
omc$time <- ifelse(omc$status %in% c(-1, 2), omc$hi, omc$lo)

fit_omc <- hazard(
  time    = omc$time,
  status  = omc$status,
  time_lower = omc$lo,
  time_upper = omc$hi,
  theta   = c(log_alpha = log(0.20), log_beta = log(1.30)),
  dist    = "loglogistic",
  fit     = TRUE,
  control = list(maxit = 400)
)

fit_omc
```
