---
title: "Example: Estimate Hazard Function (hz.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Estimate Hazard Function (hz.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `hz.*` SAS files: univariable hazard function
estimation with no covariates. The hazard function shape is fit to
time-to-event data to characterize early, late, and constant hazard phases.

---

## hz.death.AVC — Death after AVC repair

**SAS source:** `examples/hz.death.AVC.sas`

**Purpose:** Estimate the parametric hazard function for death after repair of
atrioventricular septal defects. No covariates — shape parameters only.

### SAS (original)

```sas
%HAZARD(
PROC HAZARD DATA=AVCS NOCOV NOCOR CONDITION=14;
     TIME INT_DEAD;
     EVENT DEAD;
     PARMS MUE=0.2 THALF=0.15 NU=1.4 M=1;
);
```

### R translation (stub — M2 optimizer required for fitting)

```{r hz-death-avc}
# avcs <- read.table("data/avc", header = FALSE, ...)  # load raw data

fit_hz <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = NULL,          # no covariates
  theta   = c(MUE = 0.2, THALF = 0.15, NU = 1.4, M = 1),
  dist    = "weibull",
  control = list(condition = 14, nocov = TRUE, nocor = TRUE)
)

fit_hz

# Plot hazard function (hvtiPlotR)
# hv_hazard(fit_hz)
```

---

## hz.deadp.KUL — Death after KUL procedures

**SAS source:** `examples/hz.deadp.KUL.sas`

**Purpose:** Estimate hazard function for the KUL (Katholieke Universiteit
Leuven) cardiac surgery dataset.

### R translation (stub)

```{r hz-deadp-kul}
# kul <- read.table("data/kul", ...)

fit_kul <- hazard(
  time    = kul$follow_time,
  status  = kul$dead,
  control = list(condition = 14)
)
```

---

## hz.te123.OMC / hz.tm123.OMC — Time-to-event with multiple phases (OMC)

**SAS sources:** `examples/hz.te123.OMC.sas`, `examples/hz.tm123.OMC.sas`

**Purpose:** Three-phase hazard model (early, constant, late) for the OMC
(Ontario Multicenter) dataset. `te123` uses event-time input; `tm123` uses
the same model with a time-magnitude parameterization.

### R translation (stub)

```{r hz-te123-omc}
# omc <- read.table("data/omc", ...)

fit_omc <- hazard(
  time    = omc$time,
  status  = omc$event,
  dist    = "weibull",
  control = list(phases = 3)
)
```
