---
title: "Example: Actuarial Estimates (ac.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Actuarial Estimates (ac.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(hvtiRhazard)
library(hvtiPlotR)
```

These examples correspond to the `ac.*` SAS files: nonparametric actuarial
(Kaplan-Meier style) event-time estimates used to characterize the raw data
before parametric modeling.

---

## ac.death.AVC — Actuarial estimate, AVC death

**SAS source:** `examples/ac.death.AVC.sas`

**Purpose:** Compute the Kaplan-Meier survival estimate for death after AVC
repair. This is the nonparametric foundation before fitting the parametric
hazard model.

### SAS (original)

```sas
PROC LIFETEST DATA=AVCS METHOD=KM PLOTS=(S H) OUTSURV=KMOUT;
  TIME INT_DEAD * DEAD(0);
RUN;
```

### R translation

```{r ac-death-avc}
# avcs <- read.table("data/avc", ...)

# Kaplan-Meier using survival package
library(survival)

km_fit <- survfit(Surv(INT_DEAD, DEAD) ~ 1, data = avcs)
summary(km_fit)

# Plot with hvtiPlotR
# hv_km(km_fit, time_label = "Months after repair",
#        event_label = "Freedom from death (%)")
```

### Notes

- The `ac.*` examples establish the baseline nonparametric estimate against
  which the fitted parametric hazard model is compared visually.
- In HAZARD workflow, you always run `ac.*` first to characterize the raw data,
  then `hz.*` to fit the parametric shape.
- `hvtiPlotR::hv_km()` produces publication-quality KM plots with confidence
  intervals matching the SAS `PROC LIFETEST PLOTS=(S)` output style.
