---
title: "Example: Multivariable Hazard Model (hm.*)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Multivariable Hazard Model (hm.*)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(hvtiRhazard)
library(hvtiPlotR)
```

These examples correspond to the `hm.*` SAS files: multivariable parametric
hazard modeling with covariate selection and phase-specific coefficients.

---

## hm.death.AVC — Multivariable model, death after AVC repair

**SAS source:** `examples/hm.death.AVC.sas`

**Purpose:** Final multivariable hazard model for death after repair of
atrioventricular septal defects. Covariates split across EARLY (operative
hazard) and CONSTANT (constant hazard) phases.

### SAS (original)

```sas
%HAZARD(
PROC HAZARD DATA=AVCS P CONSERVE OUTHAZ=OUTEST CONDITION=14 QUASI;
     TIME INT_DEAD;
     EVENT DEAD;
     PARMS MUE=0.3504743 THALF=0.1905077 NU=1.437416 M=1 FIXM
           MUC=4.391673E-07;
     EARLY   AGE=-0.03205774, COM_IV=1.336675, MAL=0.6872028,
             OPMOS=-0.01963377, OP_AGE=0.0002086689, STATUS=0.5169533;
     CONSTANT INC_SURG=1.375285, ORIFICE=3.11765, STATUS=1.054988;
);
```

### R translation (stub — M2 optimizer required for fitting)

```{r hm-death-avc}
# avcs <- read.table("data/avc", ...)
# avcs$INC_SURG[is.na(avcs$INC_SURG)] <- mean(avcs$INC_SURG, na.rm = TRUE)

X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                           "STATUS", "INC_SURG", "ORIFICE")])

fit_hm <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = X,
  theta   = c(
    MUE = 0.3504743, THALF = 0.1905077, NU = 1.437416, M = 1,
    MUC = 4.391673e-07,
    AGE = -0.03205774, COM_IV = 1.336675, MAL = 0.6872028,
    OPMOS = -0.01963377, OP_AGE = 0.0002086689, STATUS = 0.5169533,
    INC_SURG = 1.375285, ORIFICE = 3.11765
  ),
  dist    = "weibull",
  control = list(condition = 14, quasi = TRUE, conserve = TRUE, fix = "M")
)

fit_hm
```

---

## hm.death.AVC.deciles — Goodness of fit by decile

**SAS source:** `examples/hm.death.AVC.deciles.sas`

**Purpose:** Validate the AVC multivariable model by comparing observed vs
predicted event rates across deciles of risk.

### R translation (stub)

```{r hm-avc-deciles}
# Requires fitted model from hm.death.AVC above

# Decile goodness-of-fit plot (hvtiPlotR)
# hv_goodness_followup(fit_hm, data = avcs, group_by = "decile")
```

---

## hm.death.patient — Patient-specific hazard model

**SAS source:** `examples/hm.death.patient.sas`

**Purpose:** Patient-level parametric hazard model with individual risk scores.

### R translation (stub)

```{r hm-death-patient}
fit_patient <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = X,
  dist    = "weibull"
)
```

---

## hm.deadp.VALVES — Multivariable model, valve surgery

**SAS source:** `examples/hm.deadp.VALVES.sas`

**Purpose:** Multivariable hazard model for death after valve surgery using the
VALVES dataset.

### R translation (stub)

```{r hm-deadp-valves}
# valves <- read.table("data/valves", ...)

fit_valves <- hazard(
  time    = valves$follow_time,
  status  = valves$dead,
  x       = data.matrix(valves[, c("age", "nyha", "valve_type")]),
  dist    = "weibull"
)
```
