---
title: "Example: Hazard Prediction (hp.*)"
format: html
---

<!--
%\VignetteIndexEntry{Example: Hazard Prediction (hp.*)}
%\VignetteEngine{knitr::quarto}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| label: setup
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `hp.*` SAS files: prediction from a fitted
hazard model including survival curves, patient-level risk, and group
comparisons. These map to `predict.hazard()` and `hvtiPlotR` plotting output.

---

## hp.death.AVC — Predicted survival after AVC repair

**SAS source:** `examples/hp.death.AVC.sas`

**Purpose:** Compute and plot predicted survival and hazard curves from the
fitted AVC model.

### SAS (original)

```sas
%HAZPRED(
PROC HAZARD DATA=OUTEST IN=AVCS OUT=HPOUT;
     TIME INT_DEAD;
     PREDICT SURVIVAL HAZARD;
);
```

### R translation (partial)

```{r}
#| label: hp-death-avc
# fit_hm loaded from hm.death.AVC example

# Survival and hazard predictions on a time grid
time_grid <- seq(0, max(avcs$INT_DEAD, na.rm = TRUE), length.out = 500)

# hz_pred  <- predict(fit_hm, newdata = ..., type = "hazard")
# surv_pred <- predict(fit_hm, newdata = ..., type = "survival")

# Plot with hvtiPlotR
# hv_hazard(fit_hm, time_max = max(time_grid))
```

---

## hp.death.AVC.hm1 / hp.death.AVC.hm2 — Predictions from two models

**SAS sources:** `examples/hp.death.AVC.hm1.sas`, `examples/hp.death.AVC.hm2.sas`

**Purpose:** Compare predicted hazard curves from two fitted multivariable
models (e.g., alternative covariate specifications).

### R translation (partial)

```{r}
#| label: hp-avc-hm1-hm2
# fit_hm1, fit_hm2 — two separately fitted hazard() objects

# hv_hazard_compare(list(Model1 = fit_hm1, Model2 = fit_hm2))
```

---

## hp.death.COMPARISON — Model comparison

**SAS source:** `examples/hp.death.COMPARISON.sas`

**Purpose:** Overlay predicted survival/hazard from multiple models or covariate
scenarios on a single plot.

### R translation (partial)

```{r}
#| label: hp-death-comparison
# predict() on each fitted model, then bind and plot
# Requires: predict(fit, type = "survival")
```

---

## hp.dthip.PAIVS.time — Prediction with time-varying risk

**SAS source:** `examples/hp.dthip.PAIVS.time.sas`

**Purpose:** Predict patient-specific hazard over time for the PAIVS (Pulmonary
Atresia with Intact Ventricular Septum) dataset.

### R translation

```{r}
#| label: hp-dthip-paivs
set.seed(42)

# Synthetic PAIVS-like example for time-varying risk demonstration.
n <- 120
paivs <- data.frame(
     time = rexp(n, rate = 0.25) + 0.05,
     dead = rbinom(n, size = 1, prob = 0.65),
     vent_score = rnorm(n),
     shunt = rbinom(n, size = 1, prob = 0.4)
)

# Define piecewise windows for β(t): early, mid, late follow-up.
tv_windows <- c(1.0, 3.0)

# Weibull with time-varying coefficients via windowed design expansion.
# theta layout with two predictors and 3 windows:
# [mu, nu, vent_w1, shunt_w1, vent_w2, shunt_w2, vent_w3, shunt_w3]
fit_paivs <- hazard(
     time = paivs$time,
     status = paivs$dead,
     x = paivs[, c("vent_score", "shunt")],
     time_windows = tv_windows,
     theta = c(0.4, 1.1, 0, 0, 0, 0, 0, 0),
     dist = "weibull",
     fit = TRUE
)

# Time-specific risk summaries.
pred_df <- paivs
pred_df$lp <- predict(
     fit_paivs,
     newdata = pred_df[, c("time", "vent_score", "shunt")],
     type = "linear_predictor"
)
pred_df$haz_mult <- predict(
     fit_paivs,
     newdata = pred_df[, c("time", "vent_score", "shunt")],
     type = "hazard"
)

head(pred_df[, c("time", "lp", "haz_mult")])
```
