---
title: "Fitting Hazard Models"
subtitle: "From intercept-only to multiphase additive hazard"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
---

<!--
%\VignetteIndexEntry{Fitting Hazard Models}
%\VignetteEngine{quarto::html}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkg
)
```

```{r}
#| label: setup
library(TemporalHazard)
library(survival)
library(ggplot2)
```

This vignette demonstrates the core model-fitting workflow, progressing from
the simplest case (intercept-only, single distribution) through multivariable
models to multiphase additive hazard decomposition.  All examples use the
clinical datasets shipped with the package.


## Intercept-only model: CABG survival (KU Leuven)

The `cabgkul` dataset contains 5,880 patients who underwent primary isolated
coronary artery bypass grafting at KU Leuven between 1971 and 1987.  With only
two columns --- follow-up time and death indicator --- it is the simplest
starting point.

```{r}
#| label: kul-data
data(cabgkul)
str(cabgkul)
```

Fit an intercept-only Weibull hazard (no covariates):

```{r}
#| label: kul-weibull
fit_kul <- hazard(
  Surv(int_dead, dead) ~ 1,
  data  = cabgkul,
  dist  = "weibull",
  theta = c(mu = 0.10, nu = 1.0),
  fit   = TRUE
)

fit_kul
```

Compare the parametric fit to the Kaplan-Meier estimate:

```{r}
#| label: fig-kul-km
#| fig-cap: "Weibull parametric survival vs. Kaplan-Meier (CABG, KU Leuven)"
#| fig-width: 7
#| fig-height: 4.5
t_grid <- seq(0.01, max(cabgkul$int_dead) * 0.9, length.out = 200)
nd     <- data.frame(time = t_grid)
surv   <- predict(fit_kul, newdata = nd, type = "survival") * 100

km     <- survfit(Surv(int_dead, dead) ~ 1, data = cabgkul)
km_df  <- data.frame(time = km$time, survival = km$surv * 100)

ggplot() +
  geom_step(data = km_df, aes(time, survival, colour = "Kaplan-Meier"),
            linewidth = 0.5) +
  geom_line(data = data.frame(time = t_grid, survival = surv),
            aes(time, survival, colour = "Weibull"), linewidth = 1) +
  scale_colour_manual(values = c("Weibull" = "#0072B2",
                                 "Kaplan-Meier" = "#D55E00")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Months after CABG", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

A single Weibull captures the broad trend but misses the distinct early
operative risk and late attrition that the KM curve reveals.  This motivates
the multiphase approach below.


## Multivariable model: AVC repair

The `avc` dataset has 310 patients with 9 covariates.  We can use the formula
interface to fit a multivariable Weibull model.

```{r}
#| label: avc-data
data(avc)
avc <- na.omit(avc)
str(avc)
```

```{r}
#| label: avc-mv
fit_avc <- hazard(
  Surv(int_dead, dead) ~ age + status + mal + com_iv + inc_surg + orifice,
  data  = avc,
  dist  = "weibull",
  theta = c(mu = 0.20, nu = 1.0, rep(0, 6)),
  fit   = TRUE,
  control = list(maxit = 500)
)

fit_avc
```

The coefficient estimates give the log-hazard-ratio for each covariate.
Covariates with large positive coefficients (`mal`, `com_iv`) are associated
with higher operative risk.


## Multiphase model: additive hazard decomposition

The key innovation of the Blackstone--Naftel--Turner framework is decomposing
the hazard into additive temporal phases:

$$H(t \mid x) = \sum_{j=1}^{J} \mu_j(x) \cdot \Phi_j(t)$$

Each phase has its own temporal shape (early peaking, constant, late rising)
and its own intercept.  We specify phases with `hzr_phase()`:

```{r}
#| label: avc-mp
fit_mp <- hazard(
  Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf",      t_half = 0.5, nu = 1, m = 1),
    constant = hzr_phase("constant"),
    late     = hzr_phase("cdf",      t_half = 10,  nu = 1, m = 1)
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)

summary(fit_mp)
```

Comparing the single-phase Weibull to the multiphase fit:

```{r}
#| label: fig-avc-compare
#| fig-cap: "Single-phase Weibull vs. multiphase model against Kaplan-Meier (AVC)"
#| fig-width: 7
#| fig-height: 4.5
t_grid <- seq(0.01, max(avc$int_dead) * 0.95, length.out = 200)
nd     <- data.frame(time = t_grid)

km_avc <- survfit(Surv(int_dead, dead) ~ 1, data = avc)
km_df  <- data.frame(time = km_avc$time, survival = km_avc$surv * 100)

fit_wb <- hazard(
  Surv(int_dead, dead) ~ 1, data = avc, dist = "weibull",
  theta = c(mu = 0.20, nu = 1.0), fit = TRUE
)

surv_wb <- predict(fit_wb, newdata = nd, type = "survival") * 100
surv_mp <- predict(fit_mp, newdata = nd, type = "survival") * 100

plot_df <- rbind(
  data.frame(time = t_grid, survival = surv_wb, Model = "Single Weibull"),
  data.frame(time = t_grid, survival = surv_mp, Model = "Multiphase (3-phase)")
)

ggplot() +
  geom_step(data = km_df, aes(time, survival), colour = "grey50",
            linewidth = 0.5) +
  geom_line(data = plot_df, aes(time, survival, colour = Model),
            linewidth = 1) +
  scale_colour_manual(values = c("Single Weibull" = "#E69F00",
                                 "Multiphase (3-phase)" = "#0072B2")) +
  scale_y_continuous(limits = c(0, 100)) +
  annotate("text", x = max(t_grid) * 0.6, y = 95, label = "KM (grey)",
           size = 3, colour = "grey50") +
  labs(x = "Months after AVC repair", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The multiphase model tracks the KM curve much more closely, capturing both
the steep early decline and the gradual late attrition that a single Weibull
cannot represent.


## Multi-endpoint models: heart valve replacement

The `valves` dataset (1,533 patients) has multiple time-to-event endpoints ---
death, prosthetic valve endocarditis (PVE), and reoperation --- each with its
own follow-up time and event indicator.  The same `hazard()` call fits each
endpoint independently:

```{r}
#| label: valves-death
data(valves)
valves <- na.omit(valves)

fit_death <- hazard(
  Surv(int_dead, dead) ~ age_cop + nyha + mechvalv,
  data  = valves,
  dist  = "weibull",
  theta = c(mu = 0.10, nu = 1.0, rep(0, 3)),
  fit   = TRUE,
  control = list(maxit = 500)
)

fit_death
```

```{r}
#| label: valves-pve
fit_pve <- hazard(
  Surv(int_pve, pve) ~ age_cop + nve + mechvalv,
  data  = valves,
  dist  = "weibull",
  theta = c(mu = 0.02, nu = 1.0, rep(0, 3)),
  fit   = TRUE,
  control = list(maxit = 500)
)

fit_pve
```

Each endpoint can be modeled independently with appropriate covariates.  The
hazard model structure --- temporal shape + covariate effects --- is the same
regardless of the clinical endpoint.


## Phase types reference

The `hzr_phase()` constructor supports three temporal shapes:

| Type | Description | Typical use |
|------|-------------|-------------|
| `"cdf"` | Sigmoidal CDF shape (parameterized by `t_half`, `nu`, `m`) | Early or late phases with transient risk |
| `"constant"` | Flat hazard (no temporal shape parameters) | Ongoing background risk |
| `"g3"` | Late-phase G3 parameterization (4 parameters: `tau`, `gamma`, `alpha`, `eta`) | Late-rising risk matching C/SAS G3 output |

The `"cdf"` type covers the widest range of shapes.  Setting `t_half` small
(e.g., 0.5) creates an early-peaking phase; setting it large (e.g., 10)
creates a late-rising phase.  The `"constant"` phase needs no shape parameters.
See `vignette("mf-mathematical-foundations")` for the full mathematical
treatment.
