---
title: "Getting Started with TemporalHazard"
subtitle: "Build, fit, and predict with temporal parametric hazard models"
format:
  html:
    code-fold: false
---

<!--
%\VignetteIndexEntry{Getting Started with TemporalHazard}
%\VignetteEngine{quarto::html}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_pkg
)
```

This vignette shows the minimal workflow for fitting a parametric hazard model,
summarizing the fit, and generating predictions.

```{r}
library(TemporalHazard)
library(survival)

set.seed(1001)
n <- 180
dat <- data.frame(
  time = rexp(n, rate = 0.35) + 0.05,
  status = rbinom(n, size = 1, prob = 0.6),
  age = rnorm(n, mean = 62, sd = 11),
  nyha = sample(1:4, n, replace = TRUE),
  shock = rbinom(n, size = 1, prob = 0.18)
)

fit <- TemporalHazard::hazard(
  Surv(time, status) ~ age + nyha + shock,
  data = dat,
  theta = c(mu = 0.25, nu = 1.10, beta1 = 0, beta2 = 0, beta3 = 0),
  dist = "weibull",
  fit = TRUE,
  control = list(maxit = 300)
)

summary(fit)
```

## Prediction workflow

```{r}
new_patients <- data.frame(
  time = c(0.5, 1.5, 3.0),
  age = c(50, 65, 75),
  nyha = c(1, 3, 4),
  shock = c(0, 0, 1)
)

pred_input <- new_patients

new_patients$linear_predictor <- predict(fit, newdata = pred_input, type = "linear_predictor")
new_patients$hazard_multiplier <- predict(fit, newdata = pred_input, type = "hazard")
new_patients$survival <- predict(fit, newdata = pred_input, type = "survival")
new_patients$cumulative_hazard <- predict(fit, newdata = pred_input, type = "cumulative_hazard")

new_patients
```

## Visualizing predicted survival

The `hvtiPlotR` package provides publication-quality plotting for parametric
hazard models via `hv_hazard()` and its `plot()` method. Here we build a
survival curve over a fine time grid for a median-profile patient and overlay
the Kaplan-Meier nonparametric estimate from the raw data.

```{r}
#| label: fig-survival
#| fig-cap: "Parametric Weibull survival curve with Kaplan-Meier overlay"
#| fig-width: 7
#| fig-height: 5
if (requireNamespace("hvtiPlotR", quietly = TRUE)) {
  library(hvtiPlotR)

  # Parametric curve on a fine grid
  t_grid <- seq(0.05, max(dat$time), length.out = 80)
  curve_df <- data.frame(
    time = t_grid,
    AGE = median(dat$age),
    nyha = 2,
    shock = 0
  )
  curve_df$survival <- predict(fit, newdata = curve_df, type = "survival")

  # Kaplan-Meier empirical overlay
  km <- survival::survfit(survival::Surv(time, status) ~ 1, data = dat)
  km_df <- data.frame(
    time = km$time,
    estimate = km$surv * 100
  )

  # Build plot with hv_hazard + plot
  hz_obj <- hv_hazard(
    curve_data   = transform(curve_df, survival = survival * 100),
    x_col        = "time",
    estimate_col = "survival",
    empirical    = km_df,
    emp_x_col    = "time",
    emp_estimate_col = "estimate"
  )

  plot(hz_obj) +
    ggplot2::scale_y_continuous(limits = c(0, 100)) +
    ggplot2::labs(
      x = "Years after surgery",
      y = "Freedom from death (%)",
      title = "Parametric survival vs Kaplan-Meier"
    ) +
    hv_theme()
}
```

## Numerical helpers

The numerical helper functions remain available directly when you need stable
log-scale calculations for custom work or debugging.

```{r}
TemporalHazard::hzr_log1pexp(c(-2, 0, 2))
```
