---
title: "Example: Multivariable Hazard Model (hm.*)"
format: html
---

<!--
%\VignetteIndexEntry{Example: Multivariable Hazard Model (hm.*)}
%\VignetteEngine{quarto::html}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| label: setup
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `hm.*` SAS files: multivariable parametric
hazard modeling with covariate selection and phase-specific coefficients.

---

## hm.death.AVC — Multivariable model, death after AVC repair

**SAS source:** `examples/hm.death.AVC.sas`

**Purpose:** Final multivariable hazard model for death after repair of
atrioventricular septal defects. Covariates split across EARLY (operative
hazard) and CONSTANT (constant hazard) phases.

### SAS (original)

```sas
%HAZARD(
PROC HAZARD DATA=AVCS P CONSERVE OUTHAZ=OUTEST CONDITION=14 QUASI;
     TIME INT_DEAD;
     EVENT DEAD;
     PARMS MUE=0.3504743 THALF=0.1905077 NU=1.437416 M=1 FIXM
           MUC=4.391673E-07;
     EARLY   AGE=-0.03205774, COM_IV=1.336675, MAL=0.6872028,
             OPMOS=-0.01963377, OP_AGE=0.0002086689, STATUS=0.5169533;
     CONSTANT INC_SURG=1.375285, ORIFICE=3.11765, STATUS=1.054988;
);
```

### R translation (runnable M3 example)

```{r}
#| label: hm-death-avc
set.seed(2201)
n <- 320
avcs <- data.frame(
  INT_DEAD = rexp(n, rate = 0.6),
  AGE = rnorm(n, mean = 6, sd = 3),
  COM_IV = rbinom(n, 1, 0.2),
  MAL = rbinom(n, 1, 0.25),
  OPMOS = runif(n, 0, 18),
  OP_AGE = runif(n, 0, 12),
  STATUS = sample(1:4, n, replace = TRUE),
  INC_SURG = rbinom(n, 1, 0.15),
  ORIFICE = rbinom(n, 1, 0.1)
)

lp <- -2.4 + 0.04 * avcs$AGE + 0.55 * avcs$COM_IV + 0.45 * avcs$MAL +
  0.20 * (avcs$STATUS >= 3) + 0.35 * avcs$INC_SURG + 0.50 * avcs$ORIFICE
avcs$DEAD <- rbinom(n, 1, plogis(lp))

X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                           "STATUS", "INC_SURG", "ORIFICE")])

fit_hm <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = X,
  theta   = c(mu = 0.25, nu = 1.10, rep(0, ncol(X))),
  dist    = "weibull",
  fit     = TRUE,
  control = list(maxit = 500)
)

fit_hm
```

---

## hm.death.AVC.deciles — Goodness of fit by decile

**SAS source:** `examples/hm.death.AVC.deciles.sas`

**Purpose:** Validate the AVC multivariable model by comparing observed vs
predicted event rates across deciles of risk.

### R translation (runnable M3 example)

```{r}
#| label: hm-avc-deciles
# Decile-style risk grouping from linear predictor.
risk_score <- predict(fit_hm, type = "linear_predictor")
q_breaks <- unique(quantile(risk_score, probs = seq(0, 1, by = 0.1), na.rm = TRUE))
if (length(q_breaks) < 3) {
  decile <- cut(rank(risk_score, ties.method = "average"),
                breaks = quantile(rank(risk_score, ties.method = "average"), probs = seq(0, 1, by = 0.1)),
                include.lowest = TRUE)
} else {
  decile <- cut(risk_score, breaks = q_breaks, include.lowest = TRUE)
}

aggregate(avcs$DEAD, by = list(decile = decile), FUN = mean)
```

---

## hm.death.patient — Patient-specific hazard model

**SAS source:** `examples/hm.death.patient.sas`

**Purpose:** Patient-level parametric hazard model with individual risk scores.

### R translation (runnable M3 example)

```{r}
#| label: hm-death-patient
new_patients <- data.frame(
  time = c(0.5, 1.5, 3.0),
  AGE = c(2, 8, 12),
  COM_IV = c(0, 1, 1),
  MAL = c(0, 0, 1),
  OPMOS = c(2, 6, 10),
  OP_AGE = c(1, 4, 8),
  STATUS = c(1, 3, 4),
  INC_SURG = c(0, 0, 1),
  ORIFICE = c(0, 1, 1)
)

patient_lp <- predict(fit_hm, newdata = new_patients, type = "linear_predictor")
patient_surv <- predict(fit_hm, newdata = new_patients, type = "survival")

patient_results <- cbind(new_patients, lp = patient_lp, survival = patient_surv)
patient_results

# Plot predicted survival curves for each patient profile
t_grid <- seq(0.05, 4.0, length.out = 40)
patient_curves <- do.call(rbind, lapply(seq_len(nrow(new_patients)), function(i) {
  nd <- new_patients[rep(i, length(t_grid)), ]
  nd$time <- t_grid
  data.frame(
    time = t_grid,
    survival = predict(fit_hm, newdata = nd, type = "survival") * 100,
    patient = paste0("Patient ", i)
  )
}))

hz_patients <- hv_hazard(
  curve_data   = patient_curves,
  x_col        = "time",
  estimate_col = "survival",
  group_col    = "patient"
)

plot(hz_patients) +
  ggplot2::scale_y_continuous(limits = c(0, 100)) +
  ggplot2::labs(
    x = "Years after repair",
    y = "Freedom from death (%)",
    title = "Patient-specific predicted survival — AVC model"
  ) +
  hv_theme()
```

---

## hm.deadp.VALVES — Multivariable model, valve surgery

**SAS source:** `examples/hm.deadp.VALVES.sas`

**Purpose:** Multivariable hazard model for death after valve surgery using the
VALVES dataset.

### R translation (runnable M3 example)

```{r}
#| label: hm-deadp-valves
set.seed(2202)
n <- 260
valves <- data.frame(
  follow_time = rexp(n, rate = 0.45),
  age = rnorm(n, mean = 67, sd = 11),
  nyha = sample(1:4, n, replace = TRUE),
  valve_type = factor(sample(c("bio", "mech"), n, replace = TRUE))
)
valves$dead <- rbinom(
  n,
  1,
  plogis(-2.1 + 0.025 * valves$age + 0.35 * (valves$nyha >= 3) + 0.20 * (valves$valve_type == "mech"))
)

X_valves <- model.matrix(~ age + nyha + valve_type, data = valves)[, -1, drop = FALSE]

fit_valves <- hazard(
  time    = valves$follow_time,
  status  = valves$dead,
  x       = X_valves,
  theta   = c(mu = 0.20, nu = 1.00, rep(0, ncol(X_valves))),
  dist    = "weibull",
  fit     = TRUE,
  control = list(maxit = 400)
)

fit_valves
```
