---
title: "Inference & Diagnostics"
subtitle: "Exploratory screening, bootstrap confidence intervals, and sensitivity analysis"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
---

<!--
%\VignetteIndexEntry{Inference & Diagnostics}
%\VignetteEngine{quarto::html}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkg
)
```

```{r}
#| label: setup
library(TemporalHazard)
library(survival)
if (requireNamespace("ggplot2", quietly = TRUE)) library(ggplot2)
```

This vignette covers the diagnostic and inferential tools surrounding hazard
model fitting: exploratory covariate screening, bootstrap confidence intervals,
decile-of-risk validation, and sensitivity analysis across covariate scenarios.
These correspond to the `lg.*`, `bs.*`, and `hs.*` steps in the classic
HAZARD analysis workflow.


## Exploratory covariate screening

Before fitting the parametric hazard model, it is useful to screen candidate
covariates with simple logistic models.  This identifies transformations and
functional forms that should enter the final model.

```{r}
#| label: explore-data
data(avc)

# Candidate covariates
candidates <- c("age", "status", "mal", "com_iv", "inc_surg", "orifice")
```

Fit a univariable logistic regression for each covariate against the death
indicator:

```{r}
#| label: explore-logit
logit_fits <- lapply(candidates, function(var) {
  fmla <- as.formula(paste("dead ~", var))
  glm(fmla, data = avc, family = binomial())
})
names(logit_fits) <- candidates

# Which covariates are significant univariably?
p_values <- vapply(logit_fits, function(f) {
  coef(summary(f))[2, "Pr(>|z|)"]
}, numeric(1))

data.frame(covariate = names(p_values),
           p_value   = round(p_values, 4),
           row.names = NULL)
```

```{r}
#| label: fig-explore-age
#| fig-cap: "Exploratory: age vs. mortality with LOESS smooth"
#| fig-width: 7
#| fig-height: 4
ggplot(avc, aes(age, dead)) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = TRUE, colour = "#0072B2",
              linewidth = 0.8) +
  labs(x = "Age at operation (months)", y = "Death (0/1)") +
  theme_minimal()
```

The LOESS smooth reveals the functional form of each covariate's relationship
to mortality, guiding decisions about transformations (log, polynomial) before
the hazard model.


## Bootstrap confidence intervals

Bootstrap resampling is the primary uncertainty quantification method in the
HAZARD workflow.  Each replicate refits the model on a resampled dataset and
accumulates prediction curves; the CI is summarized across replicates.

```{r}
#| label: fit-base
# Fit the base model that bootstrapping will use
fit <- hazard(
  Surv(int_dead, dead) ~ age + status + mal + com_iv,
  data  = avc,
  dist  = "weibull",
  theta = c(mu = 0.20, nu = 1.0, rep(0, 4)),
  fit   = TRUE,
  control = list(maxit = 500)
)

# Prediction grid at a median-risk profile
t_grid <- seq(0.01, max(avc$int_dead) * 0.9, length.out = 100)
base_nd <- data.frame(time = t_grid, age = median(avc$age),
                      status = 2, mal = 0, com_iv = 0)

surv_point <- predict(fit, newdata = base_nd, type = "survival")
```

```{r}
#| label: bootstrap
set.seed(42)
n_boot <- 50  # kept small for vignette build time

boot_surv <- matrix(NA_real_, nrow = n_boot, ncol = length(t_grid))

for (b in seq_len(n_boot)) {
  idx    <- sample(nrow(avc), replace = TRUE)
  boot_d <- avc[idx, ]

  boot_fit <- tryCatch(
    hazard(
      Surv(int_dead, dead) ~ age + status + mal + com_iv,
      data    = boot_d,
      dist    = "weibull",
      theta   = coef(fit),
      fit     = TRUE,
      control = list(maxit = 300)
    ),
    error = function(e) NULL
  )

  if (!is.null(boot_fit)) {
    boot_surv[b, ] <- predict(boot_fit, newdata = base_nd,
                               type = "survival")
  }
}

# Remove failed replicates
ok <- !is.na(boot_surv[, 1])
boot_surv <- boot_surv[ok, , drop = FALSE]
message(sum(ok), " of ", n_boot, " bootstrap replicates converged")
```

```{r}
#| label: fig-bootstrap
#| fig-cap: "Parametric survival with bootstrap 95% confidence interval"
#| fig-width: 7
#| fig-height: 4.5
ci_lo <- apply(boot_surv, 2, quantile, probs = 0.025) * 100
ci_hi <- apply(boot_surv, 2, quantile, probs = 0.975) * 100

ci_df <- data.frame(time = t_grid,
                    survival = surv_point * 100,
                    lower = ci_lo, upper = ci_hi)

km     <- survfit(Surv(int_dead, dead) ~ 1, data = avc)
km_df  <- data.frame(time = km$time, survival = km$surv * 100)

ggplot() +
  geom_ribbon(data = ci_df, aes(time, ymin = lower, ymax = upper),
              fill = "grey75", alpha = 0.5) +
  geom_step(data = km_df, aes(time, survival, colour = "Kaplan-Meier"),
            linewidth = 0.5) +
  geom_line(data = ci_df, aes(time, survival, colour = "Parametric"),
            linewidth = 1) +
  scale_colour_manual(values = c("Parametric" = "#0072B2",
                                 "Kaplan-Meier" = "#D55E00")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Years after AVC repair", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## Decile-of-risk validation

A standard model diagnostic: partition patients into deciles of predicted risk
and compare observed vs. predicted event rates.  Good calibration means the
two track each other.

```{r}
#| label: fig-deciles
#| fig-cap: "Observed vs. predicted mortality by decile of risk"
#| fig-width: 7
#| fig-height: 4.5
# Linear predictor = log-relative-hazard for each patient
lp <- predict(fit, type = "linear_predictor")

# Create deciles (handle ties gracefully)
decile <- cut(rank(lp, ties.method = "average"),
              breaks = quantile(rank(lp, ties.method = "average"),
                                probs = seq(0, 1, by = 0.1)),
              include.lowest = TRUE, labels = 1:10)

cal <- aggregate(avc$dead, by = list(Decile = decile), FUN = mean)
names(cal)[2] <- "observed"
cal$Decile <- as.integer(as.character(cal$Decile))

ggplot(cal, aes(Decile, observed)) +
  geom_col(fill = "#56B4E9", alpha = 0.7) +
  geom_hline(yintercept = mean(avc$dead), linetype = "dashed",
             colour = "grey40") +
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Risk decile (1 = lowest)", y = "Observed mortality rate") +
  theme_minimal()
```


## Sensitivity analysis

Sensitivity analysis compares predictions across different covariate
configurations to assess how the model responds to risk factor changes.

```{r}
#| label: sensitivity
# Define reference and high-risk profiles
ref_profile <- data.frame(
  age    = median(avc$age),
  status = 2,
  mal    = 0,
  com_iv = 0
)

high_risk <- data.frame(
  age    = quantile(avc$age, 0.90),
  status = 4,
  mal    = 1,
  com_iv = 1
)
```

```{r}
#| label: fig-sensitivity
#| fig-cap: "Sensitivity analysis: reference vs. high-risk profile"
#| fig-width: 7
#| fig-height: 4.5
sens_curves <- do.call(rbind, lapply(
  list("Reference" = ref_profile, "High risk" = high_risk),
  function(p) {
    nd <- p[rep(1, length(t_grid)), ]
    nd$time <- t_grid
    data.frame(time = t_grid,
               survival = predict(fit, newdata = nd, type = "survival") * 100,
               Profile = deparse(substitute(p)))
  }
))
# Fix profile labels
sens_curves$Profile <- rep(c("Reference", "High risk"),
                           each = length(t_grid))
sens_curves$Profile <- factor(sens_curves$Profile,
                              levels = c("Reference", "High risk"))

ggplot(sens_curves, aes(time, survival, colour = Profile)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("Reference" = "#009E73",
                                 "High risk" = "#D55E00")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Years after AVC repair", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The gap between the curves quantifies the clinical impact of the risk factors.
The sensitivity analysis is most informative when combined with bootstrap
confidence intervals on each curve, showing whether the difference is
statistically meaningful.


## Analysis workflow summary

The complete HAZARD analysis sequence, now implemented in TemporalHazard:

1. **Exploratory screening** (`glm`) --- identify covariate transformations
   and functional forms
2. **Kaplan-Meier baseline** (`survfit`) --- nonparametric reference curve
3. **Fit hazard model** (`hazard()`) --- parametric shape + covariates
4. **Predict & visualize** (`predict()`) --- survival, hazard, risk profiles
5. **Bootstrap CIs** --- uncertainty quantification via resampling
6. **Sensitivity analysis** --- compare scenarios across covariate settings
7. **Decile validation** --- calibration check

See `vignette("getting-started")` for the minimal workflow and
`vignette("fitting-hazard-models")` for the model-fitting details.
