---
title: "Inference & Diagnostics"
subtitle: "Exploratory screening, bootstrap confidence intervals, and sensitivity analysis"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Inference & Diagnostics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkg
)
```

```{r}
#| label: setup
library(TemporalHazard)
library(survival)
library(ggplot2)
```

This vignette covers the diagnostic and inferential tools surrounding hazard
model fitting: exploratory covariate screening, bootstrap confidence intervals,
decile-of-risk validation, and sensitivity analysis across covariate scenarios.
These correspond to the `lg.*`, `bs.*`, and `hs.*` steps in the classic
HAZARD analysis workflow.


## Exploratory covariate screening

Before fitting the parametric hazard model, it is useful to screen candidate
covariates with simple logistic models.  This identifies transformations and
functional forms that should enter the final model.

```{r}
#| label: explore-data
data(avc)
avc <- na.omit(avc)

# Candidate covariates
candidates <- c("age", "status", "mal", "com_iv", "inc_surg", "orifice")
```

Fit a univariable logistic regression for each covariate against the death
indicator:

```{r}
#| label: explore-logit
logit_fits <- lapply(candidates, function(var) {
  fmla <- as.formula(paste("dead ~", var))
  glm(fmla, data = avc, family = binomial())
})
names(logit_fits) <- candidates

# Which covariates are significant univariably?
p_values <- vapply(logit_fits, function(f) {
  coef(summary(f))[2, "Pr(>|z|)"]
}, numeric(1))

data.frame(covariate = names(p_values),
           p_value   = round(p_values, 4),
           row.names = NULL)
```

```{r}
#| label: fig-explore-age
#| fig-cap: "Exploratory: age vs. mortality with LOESS smooth"
#| fig-width: 7
#| fig-height: 4
ggplot(avc, aes(age, dead)) +
  geom_jitter(height = 0.05, width = 0, alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = TRUE, colour = "#0072B2",
              linewidth = 0.8) +
  labs(x = "Age at operation (months)", y = "Death (0/1)") +
  theme_minimal()
```

The LOESS smooth reveals the functional form of each covariate's relationship
to mortality, guiding decisions about transformations (log, polynomial) before
the hazard model.


### Quantile calibration with `hzr_calibrate()`

`hzr_calibrate()` bins a continuous covariate into quantile groups and
applies the logit / Gompertz / Cox link transform, so you can eyeball
whether the covariate-outcome relationship is linear on the chosen link
scale.  It corresponds to the SAS `logit.sas` / `logitgr.sas` macros.

```{r}
#| label: calibrate-age
age_cal <- hzr_calibrate(
  x      = avc$age,
  event  = avc$dead,
  groups = 10,
  link   = "logit"
)
print(age_cal)
```

A roughly linear trend across groups means the logit link is
appropriate; strong curvature flags the need for a transform (for
example, polynomial or log).

## Nonparametric baselines

Before fitting the parametric model, run a Kaplan-Meier or Nelson
baseline.  `hzr_kaplan()` returns KM with logit-transformed confidence
limits (more accurate in the tails than Greenwood), along with interval
hazard rate, density, and restricted mean survival time.  `hzr_nelson()`
returns the Wayne Nelson cumulative hazard estimator with lognormal
confidence limits.

```{r}
#| label: km-baseline
km <- hzr_kaplan(time = avc$int_dead, status = avc$dead, conf_level = 0.95)
print(km)
```

```{r}
#| label: nelson-baseline
nel <- hzr_nelson(time = avc$int_dead, event = avc$dead, conf_level = 0.95)
print(nel)
```


## Bootstrap confidence intervals

Bootstrap resampling is the primary uncertainty quantification method in the
HAZARD workflow.  Each replicate refits the model on a resampled dataset and
accumulates prediction curves; the CI is summarized across replicates.

```{r}
#| label: fit-base
# Fit the base model that bootstrapping will use
fit <- hazard(
  Surv(int_dead, dead) ~ age + status + mal + com_iv,
  data  = avc,
  dist  = "weibull",
  theta = c(mu = 0.20, nu = 1.0, rep(0, 4)),
  fit   = TRUE,
  control = list(maxit = 500)
)

# Prediction grid at a median-risk profile
t_grid <- seq(0.01, max(avc$int_dead) * 0.9, length.out = 100)
base_nd <- data.frame(time = t_grid, age = median(avc$age),
                      status = 2, mal = 0, com_iv = 0)

surv_point <- predict(fit, newdata = base_nd, type = "survival")
```

`hzr_bootstrap()` wraps the resample-and-refit loop: it returns a long
data frame of per-replicate coefficient estimates (`$replicates`) and a
parameter-level summary (mean, SD, 95 % percentile CI) in `$summary`.
It corresponds to the SAS `bootstrap.hazard.sas` and
`bootstrap.summary.sas` macros.

```{r}
#| label: bootstrap
set.seed(42)
boot <- hzr_bootstrap(fit, n_boot = 30)  # kept small for vignette build time
print(boot)
```

The parameter-level CIs above come directly from the bootstrap
distribution.  Prediction-level CIs (on the survival curve or hazard
profile) require pivoting `boot$replicates` to wide format and
evaluating the parametric survival formula once per replicate.  That
workflow is out of scope for this section, but the building blocks are
all in the returned object.


## Goodness-of-fit overlay

`hzr_gof()` computes parametric predictions against the nonparametric
Kaplan-Meier estimator and the Conservation-of-Events observed-vs-
expected event counts.  It corresponds to the SAS `hazplot.sas` macro.

```{r}
#| label: gof-summary
gof <- hzr_gof(fit)
print(gof)
```

The observed vs. expected ratio (printed at the bottom) is the
Conservation-of-Events check: a well-specified model recovers the event
count exactly, so a ratio close to 1 is what we want.  Ratios well
above or below 1 flag either a misspecified shape or a covariate that
isn't really linear on the log-hazard scale.


## Decile-of-risk calibration

`hzr_deciles()` partitions patients into deciles of predicted risk at a
chosen time point and compares observed with expected event counts
decile-by-decile, returning a chi-square goodness-of-fit test.  It
corresponds to the SAS `deciles.hazard.sas` macro.

```{r}
#| label: deciles
deciles <- hzr_deciles(fit, time = 120, groups = 10)
print(deciles)
```

The chi-square test statistic and p-value summarise whether the
observed decile-by-decile event counts are consistent with what the
parametric model predicts.  A good visual check is to plot the two
columns side by side:

```{r}
#| label: fig-deciles
#| fig-cap: "Observed vs. expected mortality by decile of risk"
#| fig-width: 7
#| fig-height: 4.5
decile_df <- as.data.frame(deciles)
cal_long <- rbind(
  data.frame(group = decile_df$group,
             rate  = decile_df$observed_rate,
             Series = "Observed"),
  data.frame(group = decile_df$group,
             rate  = decile_df$expected_rate,
             Series = "Expected")
)
ggplot(cal_long, aes(group, rate, fill = Series)) +
  geom_col(position = position_dodge(width = 0.7), alpha = 0.8) +
  scale_fill_manual(values = c("Observed" = "#56B4E9",
                                "Expected" = "#E69F00")) +
  scale_x_continuous(breaks = seq_len(nrow(decile_df))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Risk decile (1 = lowest)", y = "Mortality rate", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## Sensitivity analysis

Sensitivity analysis compares predictions across different covariate
configurations to assess how the model responds to risk factor changes.

```{r}
#| label: sensitivity
# Define reference and high-risk profiles
ref_profile <- data.frame(
  age    = median(avc$age),
  status = 2,
  mal    = 0,
  com_iv = 0
)

high_risk <- data.frame(
  age    = quantile(avc$age, 0.90),
  status = 4,
  mal    = 1,
  com_iv = 1
)
```

```{r}
#| label: fig-sensitivity
#| fig-cap: "Sensitivity analysis: reference vs. high-risk profile"
#| fig-width: 7
#| fig-height: 4.5
sens_curves <- do.call(rbind, lapply(
  list("Reference" = ref_profile, "High risk" = high_risk),
  function(p) {
    nd <- p[rep(1, length(t_grid)), ]
    nd$time <- t_grid
    data.frame(time = t_grid,
               survival = predict(fit, newdata = nd, type = "survival") * 100,
               Profile = deparse(substitute(p)))
  }
))
# Fix profile labels
sens_curves$Profile <- rep(c("Reference", "High risk"),
                           each = length(t_grid))
sens_curves$Profile <- factor(sens_curves$Profile,
                              levels = c("Reference", "High risk"))

ggplot(sens_curves, aes(time, survival, colour = Profile)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("Reference" = "#009E73",
                                 "High risk" = "#D55E00")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Months after AVC repair", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The gap between the curves quantifies the clinical impact of the risk factors.
The sensitivity analysis is most informative when combined with bootstrap
confidence intervals on each curve, showing whether the difference is
statistically meaningful.


## Analysis workflow summary

The complete HAZARD analysis sequence, now implemented in TemporalHazard:

1. **Exploratory screening** (`glm`, `hzr_calibrate()`) --- identify
   covariate transformations and functional forms
2. **Nonparametric baselines** (`hzr_kaplan()`, `hzr_nelson()`) ---
   reference KM / Nelson cumulative hazard estimators
3. **Fit hazard model** (`hazard()`) --- parametric shape + covariates
4. **Predict & visualize** (`predict()`) --- survival, hazard, risk profiles
5. **Goodness-of-fit overlay** (`hzr_gof()`) --- parametric vs. KM +
   Conservation-of-Events check
6. **Bootstrap CIs** (`hzr_bootstrap()`) --- uncertainty quantification
   via resampling
7. **Sensitivity analysis** (`predict()` on covariate profiles) ---
   compare scenarios across risk factors
8. **Decile calibration** (`hzr_deciles()`) --- chi-square observed
   vs. expected by risk decile

See `vignette("getting-started")` for the minimal workflow and
`vignette("fitting-hazard-models")` for the model-fitting details.
