---
title: "Example: Logistic/Gompertz Exploratory Plots (lg.*)"
format: html
---

<!--
%\VignetteIndexEntry{Example: Logistic/Gompertz Exploratory Plots (lg.*)}
%\VignetteEngine{quarto::html}
%\VignetteEncoding{UTF-8}
-->

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r}
#| label: setup
library(TemporalHazard)
library(hvtiPlotR)
```

These examples correspond to the `lg.*` SAS files: exploratory logistic and
Gompertz plots used to identify covariate transformations and preliminary
functional forms *before* fitting the parametric hazard model. These are
diagnostic/exploratory tools, not the primary analysis.

---

## lg.death.AVC — Logistic/Gompertz plots, AVC death

**SAS source:** `examples/lg.death.AVC.sas`

**Purpose:** For each candidate covariate, fit a logistic (early hazard) or
Gompertz (late hazard) model and plot the estimated relationship. This guides
covariate transformation decisions (log, power, inverse) before the hazard
model.

### SAS (original)

```sas
%LOGIT(
  DATA  = AVCS,
  EVENT = DEAD,
  VARS  = AGE STATUS INC_SURG OPMOS MAL COM_IV ORIFICE OP_AGE
);
```

### R translation (runnable M3 example)

```{r}
#| label: lg-death-avc
set.seed(5201)
n <- 360
avcs <- data.frame(
  INT_DEAD = rexp(n, rate = 0.6),
  AGE = rnorm(n, mean = 6, sd = 3),
  STATUS = sample(1:4, n, replace = TRUE),
  INC_SURG = rbinom(n, 1, 0.15),
  OPMOS = runif(n, 0, 18),
  MAL = rbinom(n, 1, 0.25),
  COM_IV = rbinom(n, 1, 0.2),
  ORIFICE = rbinom(n, 1, 0.1),
  OP_AGE = runif(n, 0, 12)
)

lp <- -2.7 + 0.05 * avcs$AGE + 0.20 * avcs$STATUS + 0.55 * avcs$INC_SURG +
  0.03 * avcs$OPMOS + 0.45 * avcs$MAL + 0.60 * avcs$COM_IV +
  0.35 * avcs$ORIFICE + 0.01 * avcs$OP_AGE
avcs$DEAD <- rbinom(n, 1, plogis(lp))

candidates <- c("AGE", "STATUS", "INC_SURG", "OPMOS",
                "MAL", "COM_IV", "ORIFICE", "OP_AGE")

# Fit logistic model for each candidate covariate
logit_fits <- lapply(candidates, function(var) {
  fmla <- as.formula(paste("DEAD ~", var))
  glm(fmla, data = avcs, family = binomial())
})
names(logit_fits) <- candidates

# Inspect: which covariates show monotone relationships?
coef_tables <- lapply(logit_fits, function(f) coef(summary(f)))
coef_tables[["AGE"]]

# Plot univariable covariate effects using hvtiPlotR EDA
hv_obj <- hv_eda(avcs, x = "AGE", y = "DEAD")
plot(hv_obj) +
  ggplot2::labs(
    x = "Age at operation (years)",
    y = "Death",
    title = "Exploratory: AGE vs mortality"
  ) +
  hv_theme()
```

---

## lg.dead30d.AVC — 30-day mortality logistic model

**SAS source:** `examples/lg.dead30d.AVC.sas`

**Purpose:** Logistic regression for early (30-day) mortality after AVC repair.
Identifies covariates driving operative risk.

### R translation (runnable M3 example)

```{r}
#| label: lg-dead30d-avc
# Define 30-day event indicator
avcs$dead30 <- as.integer(avcs$DEAD == 1 & avcs$INT_DEAD <= 1)

fit_30d <- glm(dead30 ~ AGE + STATUS + COM_IV + MAL,
               data = avcs, family = binomial())
summary(fit_30d)
```

---

## lg.deadlate.AVC — Late mortality logistic model

**SAS source:** `examples/lg.deadlate.AVC.sas`

**Purpose:** Logistic regression restricted to patients surviving 30 days.
Identifies covariates driving late (constant-phase) hazard.

### R translation (runnable M3 example)

```{r}
#| label: lg-deadlate-avc
# Restrict to survivors of operative period
avcs_late <- avcs[avcs$INT_DEAD > 1 | avcs$DEAD == 0, ]
avcs_late$dead_late <- as.integer(avcs_late$DEAD == 1)

fit_late <- glm(dead_late ~ INC_SURG + ORIFICE + STATUS,
                data = avcs_late, family = binomial())
summary(fit_late)
```

### Note on lg.* → hz.* workflow

The typical HAZARD analysis sequence is:

1. `lg.*` — identify covariate transformations and functional forms
2. `ac.*` — nonparametric estimate for baseline characterization
3. `hz.*` — fit parametric hazard function shape
4. `hm.*` — stepwise multivariable model
5. `hp.*` — predictions from final model
6. `bs.*` — bootstrap confidence intervals
7. `hs.*` — sensitivity and scenario analyses
