---
title: "Migrating from SAS HAZARD to TemporalHazard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from SAS HAZARD to TemporalHazard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(TemporalHazard)
```

## Overview

The legacy HAZARD program is a SAS macro (`%HAZARD(...)`) that wraps a compiled
C executable. It models event-time data using a parametric hazard function with
multiple phases. This vignette is the canonical reference for translating a SAS
HAZARD analysis into native R using `TemporalHazard`.

The full formal argument mapping table is available programmatically:

```{r mapping-table}
knitr::kable(
  hzr_argument_mapping(),
  caption = "Formal argument map: SAS HAZARD/C → hazard()",
  col.names = c(
    "SAS Statement", "Legacy Input", "C Concept",
    "R Parameter", "Required", "Expected Type",
    "Transform Rule", "Status", "Notes"
  )
)
```

---

## Statement-by-statement mapping

### `PROC HAZARD DATA=`

```sas
PROC HAZARD DATA=AVCS NOCOV NOCOR CONDITION=14;
```

Maps to `hazard()` `control` list:

```{r, eval=FALSE}
fit <- hazard(
  ...,
  control = list(
    nocov      = TRUE,   # suppress covariance output
    nocor      = TRUE,   # suppress correlation output
    condition  = 14      # CONDITION= switch
  )
)
```

Additional `PROC HAZARD` options with no direct R equivalent yet are passed
through `...` as named arguments and stored in `fit$legacy_args`.

---

### `TIME`

```sas
TIME INT_DEAD;
```

Maps directly to `time`:

```{r, eval=FALSE}
fit <- hazard(
  time   = avcs$INT_DEAD,
  ...
)
```

- Must be a non-negative numeric vector in the same time unit as the original
  analysis (months, in the AVC example).
- Missing values are not permitted.

---

### `EVENT`

```sas
EVENT DEAD;
```

Maps to `status`:

```{r, eval=FALSE}
fit <- hazard(
  status = avcs$DEAD,   # 1 = event, 0 = censored
  ...
)
```

- `TemporalHazard` coerces `status` to `numeric`; logical vectors are also accepted.
- The HAZARD convention uses `1` = occurred, `0` = censored/alive at last contact.

---

### `PARMS`

```sas
PARMS MUE=0.3504743 THALF=0.1905077 NU=1.437416 M=1 FIXM
      MUC=4.391673E-07;
```

Maps to `theta` (coefficient/parameter vector) and `control` (fix flags):

```{r, eval=FALSE}
fit <- hazard(
  theta   = c(MUE = 0.3504743, THALF = 0.1905077, NU = 1.437416,
              M = 1,           MUC   = 4.391673e-07),
  control = list(
    fix = c("M")   # FIXM → freeze M during optimization
  ),
  ...
)
```

> **Note:** Full `PARMS` keyword parsing (FIX flags, phase selectors) is planned
> for M2. For now supply a plain numeric vector via `theta` and record any FIX
> constraints in `control$fix`.

---

### `EARLY` and `CONSTANT` covariate blocks

```sas
EARLY  AGE=-0.03205774, COM_IV=1.336675, MAL=0.6872028,
       OPMOS=-0.01963377, OP_AGE=0.0002086689, STATUS=0.5169533;
CONSTANT INC_SURG=1.375285, ORIFICE=3.11765, STATUS=1.054988;
```

In HAZARD, `EARLY` and `CONSTANT` define phase-specific covariate coefficients.
In `TemporalHazard` these are unified into a single design matrix `x` and
coefficient vector `theta` during M1. Phase assignment will be formalised in
M2 when the multi-phase likelihood is implemented.

**Current convention (M1):** combine all covariates into `x` and supply the
corresponding starting coefficients in `theta`.

```{r, eval=FALSE}
# Build design matrix from the AVC data set
X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                           "STATUS", "INC_SURG", "ORIFICE")])

# Starting values from SAS EARLY + CONSTANT blocks combined
theta_start <- c(
  AGE      = -0.03205774,
  COM_IV   =  1.336675,
  MAL      =  0.6872028,
  OPMOS    = -0.01963377,
  OP_AGE   =  0.0002086689,
  STATUS   =  0.5169533,   # EARLY phase coefficient
  INC_SURG =  1.375285,
  ORIFICE  =  3.11765
)

fit <- hazard(
  time   = avcs$INT_DEAD,
  status = avcs$DEAD,
  x      = X,
  theta  = theta_start,
  dist   = "weibull"
)
```

---

### `SELECTION`

```sas
SELECTION SLE=0.2 SLS=0.1;
```

Stepwise variable selection is **not yet implemented** in `TemporalHazard`. Planned
for M3. During migration, resolve your final variable set in the SAS analysis
first, then port the confirmed model to R.

---

## Full worked example: AVC death after repair

This mirrors the final multivariable model from
`examples/hm.death.AVC.sas` in the reference C repository.

### SAS (original)

```sas
%HAZARD(
PROC HAZARD DATA=AVCS P CONSERVE OUTHAZ=OUTEST CONDITION=14 QUASI;
     TIME INT_DEAD;
     EVENT DEAD;
     PARMS MUE=0.3504743 THALF=0.1905077 NU=1.437416 M=1 FIXM
           MUC=4.391673E-07;
     EARLY   AGE=-0.03205774, COM_IV=1.336675,  MAL=0.6872028,
             OPMOS=-0.01963377, OP_AGE=0.0002086689, STATUS=0.5169533;
     CONSTANT INC_SURG=1.375285, ORIFICE=3.11765, STATUS=1.054988;
);
```

### R equivalent (M1 stub — stores inputs, optimization pending)

```{r avc-example, eval=FALSE}
# Assumed: avcs is a data.frame read from the AVC flat file
# (same variables as the SAS DATA step)

avcs <- avcs |>
  transform(
    LN_AGE   = log(AGE),
    LN_OPMOS = log(OPMOS),
    LN_INC   = ifelse(is.na(INC_SURG), NA, log(INC_SURG + 1)),
    LN_NYHA  = log(STATUS)
  )

# Replace missing INC_SURG with column mean (mirrors PROC STANDARD REPLACE)
avcs$INC_SURG[is.na(avcs$INC_SURG)] <- mean(avcs$INC_SURG, na.rm = TRUE)

X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                           "STATUS", "INC_SURG", "ORIFICE")])

fit <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = X,
  theta   = c(
    # Hazard shape parameters (from PARMS)
    MUE   = 0.3504743,
    THALF = 0.1905077,
    NU    = 1.437416,
    M     = 1,
    MUC   = 4.391673e-07,
    # Covariate coefficients (from EARLY + CONSTANT blocks)
    AGE      = -0.03205774,
    COM_IV   =  1.336675,
    MAL      =  0.6872028,
    OPMOS    = -0.01963377,
    OP_AGE   =  0.0002086689,
    STATUS   =  0.5169533,
    INC_SURG =  1.375285,
    ORIFICE  =  3.11765
  ),
  dist    = "weibull",
  control = list(
    condition = 14,
    quasi     = TRUE,
    conserve  = TRUE,
    fix       = c("M")   # FIXM
  )
)

fit
```

---

## Data preparation differences

| SAS HAZARD                             | TemporalHazard R                                     |
|----------------------------------------|---------------------------------------------------|
| `PROC STANDARD REPLACE` for missing    | Replace with `mean(..., na.rm = TRUE)` manually  |
| Log transforms in `DATA` step          | `transform()` or `dplyr::mutate()`                |
| Variable labels via `LABEL`            | Column names of `x` matrix carry through          |
| `FILENAME` / `INFILE` / `INPUT`       | `read.table()`, `read.csv()`, or `haven::read_sas()` |
| SAS formats (date, currency, etc.)     | Standard R numeric; apply `as.Date()` if needed  |

---

## Output object

`hazard()` returns a list of class `hazard`:

| Slot                  | Contents                                                     |
|-----------------------|--------------------------------------------------------------|
| `$call`               | Unevaluated `match.call()` for reproducibility              |
| `$spec$dist`          | Baseline distribution label                                 |
| `$spec$control`       | Control options passed at construction                      |
| `$data$time`          | Follow-up time vector                                       |
| `$data$status`        | Event indicator (numeric)                                   |
| `$data$x`             | Design matrix                                               |
| `$fit$theta`          | Coefficient vector (starting values at M1; fitted at M2+)  |
| `$fit$converged`      | `NA` at M1; `TRUE`/`FALSE` from M2 optimizer               |
| `$fit$objective`      | Log-likelihood at convergence (M2+)                        |
| `$legacy_args`        | Named pass-through arguments for parity                     |

---

## Prediction

`predict.hazard()` currently supports two output types:

```{r predict-example, eval=FALSE}
# Linear predictor (X %*% theta)
eta <- predict(fit, type = "linear_predictor")

# Hazard scale (exp(eta))
hz  <- predict(fit, type = "hazard")
```

From M2 onward, `"survival"` and `"cumulative_hazard"` types will be added to
mirror the `P` (predict/print) option in `PROC HAZARD`.

---

## Rcpp acceleration note

`TemporalHazard` is currently pure R. If profiling against large real datasets
reveals bottlenecks in the likelihood kernel or optimizer inner loop, those
specific functions will be re-implemented with Rcpp. The public interface
(`hazard()`, `predict.hazard()`, etc.) will not change.
