---
title: "Migrating from SAS HAZARD to TemporalHazard"
format: html
vignette: >
  %\VignetteIndexEntry{Migrating from SAS HAZARD to TemporalHazard}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_pkg
)
```

```{r}
#| label: setup
library(TemporalHazard)
```

## Overview

The legacy HAZARD program is a SAS macro (`%HAZARD(...)`) that wraps a compiled
C executable implementing the multiphase parametric hazard model of Blackstone,
Naftel, and Turner (1986). This vignette is the canonical reference for
translating a SAS HAZARD analysis into native R using `TemporalHazard`.

The full formal argument mapping table is available programmatically:

```{r}
#| label: mapping-table
knitr::kable(
  TemporalHazard::hzr_argument_mapping(),
  caption = "Formal argument map: SAS HAZARD/C → hazard()",
  col.names = c(
    "SAS Statement", "Legacy Input", "C Concept",
    "R Parameter", "Required", "Expected Type",
    "Transform Rule", "Status", "Notes"
  )
)
```

---

## Statement-by-statement mapping

### `PROC HAZARD DATA=`

```sas
PROC HAZARD DATA=AVCS NOCOV NOCOR CONDITION=14;
```

Maps to `hazard()` `control` list:

```{r}
#| eval: false
fit <- hazard(
  ...,
  control = list(
    nocov      = TRUE,   # suppress covariance output
    nocor      = TRUE,   # suppress correlation output
    condition  = 14      # CONDITION= switch
  )
)
```

Additional `PROC HAZARD` options with no direct R equivalent yet are passed
through `...` as named arguments and stored in `fit$legacy_args`.

---

### `TIME`

```sas
TIME INT_DEAD;
```

Maps directly to `time`:

```{r}
#| eval: false
fit <- hazard(
  time   = avcs$INT_DEAD,
  ...
)
```

- Must be a non-negative numeric vector in the same time unit as the original
  analysis (months, in the AVC example).
- Missing values are not permitted.

---

### `EVENT`

```sas
EVENT DEAD;
```

Maps to `status`:

```{r}
#| eval: false
fit <- hazard(
  status = avcs$DEAD,   # 1 = event, 0 = censored
  ...
)
```

- `TemporalHazard` coerces `status` to `numeric`; logical vectors are also accepted.
- The HAZARD convention uses `1` = occurred, `0` = censored/alive at last contact.

---

### `PARMS`

```sas
PARMS MUE=0.3504743 THALF=0.1905077 NU=1.437416 M=1 FIXM
      MUC=4.391673E-07;
```

Maps to `theta` (coefficient/parameter vector) and `control` (fix flags):

```{r}
#| eval: false
fit <- hazard(
  theta   = c(MUE = 0.3504743, THALF = 0.1905077, NU = 1.437416,
              M = 1,           MUC   = 4.391673e-07),
  control = list(
    fix = c("M")   # FIXM → freeze M during optimization
  ),
  ...
)
```

> **Note:** Full SAS `PARMS` syntax is not mirrored one-for-one in the public API.
> In the current package, supply the parameter vector directly via `theta` and
> record fixed-parameter intent in `control$fix`. The legacy parity helpers do
> generate SAS-style control text when comparing against the historical binaries.

---

### `EARLY` and `CONSTANT` covariate blocks

```sas
EARLY  AGE=-0.03205774, COM_IV=1.336675, MAL=0.6872028,
       OPMOS=-0.01963377, OP_AGE=0.0002086689, STATUS=0.5169533;
CONSTANT INC_SURG=1.375285, ORIFICE=3.11765, STATUS=1.054988;
```

In HAZARD, `EARLY` and `CONSTANT` define phase-specific covariate coefficients.
In `TemporalHazard` these are unified into a single design matrix `x` and
coefficient vector `theta` during M1. Phase assignment will be formalised in
M2 when the multi-phase likelihood is implemented.

**Current convention (M1):** combine all covariates into `x` and supply the
corresponding starting coefficients in `theta`.

```{r}
#| eval: false
# Build design matrix from the AVC data set
X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                           "STATUS", "INC_SURG", "ORIFICE")])

# Starting values from SAS EARLY + CONSTANT blocks combined
theta_start <- c(
  AGE      = -0.03205774,
  COM_IV   =  1.336675,
  MAL      =  0.6872028,
  OPMOS    = -0.01963377,
  OP_AGE   =  0.0002086689,
  STATUS   =  0.5169533,   # EARLY phase coefficient
  INC_SURG =  1.375285,
  ORIFICE  =  3.11765
)

fit <- hazard(
  time   = avcs$INT_DEAD,
  status = avcs$DEAD,
  x      = X,
  theta  = theta_start,
  dist   = "weibull"
)
```

---

### `SELECTION`

```sas
SELECTION SLE=0.2 SLS=0.1;
```

`TemporalHazard` now ships `hzr_stepwise()`, which implements forward,
backward, and two-way stepwise selection with SAS-style SLENTRY /
SLSTAY thresholds, phase-specific entry for multiphase models, and a
MOVE oscillation guard.  FAST-screening (Lawless-Singhal approximate
Wald updates) is not yet implemented; every candidate currently gets a
full refit.  See `?hzr_stepwise` for the full option list.

---

## SAS macro equivalents

The legacy SAS distribution ships a suite of macros for nonparametric
baselines, goodness-of-fit diagnostics, bootstrap inference, and
variable calibration that sit alongside `PROC HAZARD`.  Each has an R
equivalent in `TemporalHazard`:

| SAS macro | R function | Purpose |
|:---|:---|:---|
| `kaplan.sas` | `hzr_kaplan()` | KM survival with logit-transformed exact CL |
| `nelsonl.sas` / `nelsont.sas` | `hzr_nelson()` | Nelson cumulative hazard with lognormal CL |
| `hazplot.sas` | `hzr_gof()` | Parametric vs. KM overlay + CoE goodness-of-fit |
| `deciles.hazard.sas` | `hzr_deciles()` | Decile-of-risk calibration + chi-square GOF |
| `logit.sas` / `logitgr.sas` | `hzr_calibrate()` | Quantile-group calibration (logit / Gompertz / Cox link) |
| `bootstrap.hazard.sas` | `hzr_bootstrap()` | Bootstrap resampling + summary |
| `markov.sas` | `hzr_competing_risks()` | Aalen-Johansen competing-risks incidence |

Minimal illustrations on the AVC dataset follow.  The
`vignette("inference-diagnostics")` walkthrough builds these into a
full analysis workflow.

```{r}
#| label: macro-demos-setup
library(survival)
data(avc)
avc <- na.omit(avc)

# Base parametric fit for downstream diagnostics
fit <- hazard(
  Surv(int_dead, dead) ~ age + status + mal + com_iv,
  data = avc, dist = "weibull",
  theta = c(mu = 0.2, nu = 1, rep(0, 4)),
  fit = TRUE, control = list(maxit = 500)
)
```

```{r}
#| label: kaplan-demo
km <- hzr_kaplan(time = avc$int_dead, status = avc$dead)
head(km, 4)
```

```{r}
#| label: nelson-demo
nel <- hzr_nelson(time = avc$int_dead, event = avc$dead)
head(nel, 4)
```

```{r}
#| label: gof-demo
head(hzr_gof(fit), 4)
```

```{r}
#| label: deciles-demo
hzr_deciles(fit, time = 120, groups = 10)
```

```{r}
#| label: calibrate-demo
hzr_calibrate(avc$age, avc$dead, groups = 10, link = "logit")
```

```{r}
#| label: bootstrap-demo
set.seed(1)
hzr_bootstrap(fit, n_boot = 20)  # small for vignette build
```

```{r}
#| label: competing-demo
data(valves)
# Synthesize a competing-risks indicator for illustration.
valves$ev <- with(valves, ifelse(dead == 1, 1L,
                                   ifelse(pve == 1,  2L,
                                   ifelse(reop == 1, 3L, 0L))))
head(hzr_competing_risks(valves$int_dead, valves$ev), 4)
```

---

## Full worked example: AVC death after repair

This mirrors the final multivariable model from
`examples/hm.death.AVC.sas` in the reference C repository.

### SAS (original)

```sas
%HAZARD(
PROC HAZARD DATA=AVCS P CONSERVE OUTHAZ=OUTEST CONDITION=14 QUASI;
     TIME INT_DEAD;
     EVENT DEAD;
     PARMS MUE=0.3504743 THALF=0.1905077 NU=1.437416 M=1 FIXM
           MUC=4.391673E-07;
     EARLY   AGE=-0.03205774, COM_IV=1.336675,  MAL=0.6872028,
             OPMOS=-0.01963377, OP_AGE=0.0002086689, STATUS=0.5169533;
     CONSTANT INC_SURG=1.375285, ORIFICE=3.11765, STATUS=1.054988;
);
```

### R equivalent (current runnable translation pattern)

```{r}
#| label: avc-example
#| eval: false
# Assumed: avcs is a data.frame read from the AVC flat file
# (same variables as the SAS DATA step)

avcs <- avcs |>
  transform(
    LN_AGE   = log(AGE),
    LN_OPMOS = log(OPMOS),
    LN_INC   = ifelse(is.na(INC_SURG), NA, log(INC_SURG + 1)),
    LN_NYHA  = log(STATUS)
  )

# Replace missing INC_SURG with column mean (mirrors PROC STANDARD REPLACE)
avcs$INC_SURG[is.na(avcs$INC_SURG)] <- mean(avcs$INC_SURG, na.rm = TRUE)

X <- data.matrix(avcs[, c("AGE", "COM_IV", "MAL", "OPMOS", "OP_AGE",
                           "STATUS", "INC_SURG", "ORIFICE")])

fit <- hazard(
  time    = avcs$INT_DEAD,
  status  = avcs$DEAD,
  x       = X,
  theta   = c(
    # Hazard shape parameters (from PARMS)
    MUE   = 0.3504743,
    THALF = 0.1905077,
    NU    = 1.437416,
    M     = 1,
    MUC   = 4.391673e-07,
    # Covariate coefficients (from EARLY + CONSTANT blocks)
    AGE      = -0.03205774,
    COM_IV   =  1.336675,
    MAL      =  0.6872028,
    OPMOS    = -0.01963377,
    OP_AGE   =  0.0002086689,
    STATUS   =  0.5169533,
    INC_SURG =  1.375285,
    ORIFICE  =  3.11765
  ),
  dist    = "weibull",
  control = list(
    condition = 14,
    quasi     = TRUE,
    conserve  = TRUE,
    fix       = c("M")   # FIXM
  )
)

fit
```

---

## Data preparation differences

| SAS HAZARD                             | TemporalHazard R                                     |
|----------------------------------------|---------------------------------------------------|
| `PROC STANDARD REPLACE` for missing    | Replace with `mean(..., na.rm = TRUE)` manually  |
| Log transforms in `DATA` step          | `transform()` or `dplyr::mutate()`                |
| Variable labels via `LABEL`            | Column names of `x` matrix carry through          |
| `FILENAME` / `INFILE` / `INPUT`       | `read.table()`, `read.csv()`, or `haven::read_sas()` |
| SAS formats (date, currency, etc.)     | Standard R numeric; apply `as.Date()` if needed  |

---

## Output object

`hazard()` returns a list of class `hazard`:

| Slot                  | Contents                                                     |
|-----------------------|--------------------------------------------------------------|
| `$call`               | Unevaluated `match.call()` for reproducibility              |
| `$spec$dist`          | Baseline distribution label                                 |
| `$spec$control`       | Control options passed at construction                      |
| `$data$time`          | Follow-up time vector                                       |
| `$data$status`        | Event indicator (numeric)                                   |
| `$data$x`             | Design matrix                                               |
| `$fit$theta`          | Coefficient vector (starting values at M1; fitted at M2+)  |
| `$fit$converged`      | `NA` at M1; `TRUE`/`FALSE` from M2 optimizer               |
| `$fit$objective`      | Log-likelihood at convergence (M2+)                        |
| `$legacy_args`        | Named pass-through arguments for parity                     |

---

## Prediction

`predict.hazard()` currently supports two output types:

```{r}
#| eval: false
#| label: predict-example
# Linear predictor (X %*% theta)
eta <- predict(fit, type = "linear_predictor")

# Hazard scale (exp(eta))
hz  <- predict(fit, type = "hazard")
```

`"survival"` and `"cumulative_hazard"` types are also supported, mirroring
the `P` (predict/print) option in `PROC HAZARD`.  Pass `decompose = TRUE`
for multiphase fits to get per-phase contributions.

---

## Known limitations vs. SAS HAZARD

A SAS HAZARD veteran migrating to `TemporalHazard` should be aware of
the following scope limits as of v0.9.8.  Detailed status per feature is
tracked in `inst/dev/SAS-PARITY-GAP-ANALYSIS.md` and
`inst/dev/DEVELOPMENT-PLAN.md`.

### Stepwise variable selection (`SELECTION` statement)

- **Supported:** `hzr_stepwise()` implements forward / backward /
  two-way stepwise with SAS-style SLENTRY / SLSTAY thresholds,
  per-phase entry for multiphase, and a MOVE oscillation guard.
- **Not yet supported:** FAST-screening (Lawless-Singhal approximate
  Wald updates).  Every candidate currently gets a full refit, which
  is slower but always correct.

### Per-phase time-varying windows

- **Supported:** `time_windows` applies one piecewise-constant cut-point
  set across all covariates.
- **Not yet supported:** distinct `EARLY`/`LATE` window sets per phase.
  Workaround: expand the design matrix manually before calling
  `hazard()`.

### Output datasets (`OUTEST=`, `OUTVCOV=`)

- The R equivalents are `coef(fit)` and `vcov(fit)` in memory; there
  is no automatic dataset-export mode.  `saveRDS()` them yourself if
  you need an on-disk artefact.

### Density / quantile prediction types

- `predict.hazard()` covers `hazard`, `cumulative_hazard`, `survival`,
  and `linear_predictor`.  Density and quantile (median survival)
  types are not wired up; derive them from `cumulative_hazard` and
  `survival` predictions if needed.

### Previously listed gaps that have since been closed

For users migrating from older TemporalHazard versions or reading
older SAS parity notes:

- **`weights` on all distributions** — shipped v0.9.6.  Weibull,
  exponential, log-logistic, log-normal, and multiphase all honour
  observation weights end-to-end (LL + analytic gradient + multiphase
  Conservation of Events).
- **`Surv(start, stop, event)` with `start > 0`** — shipped v0.9.7.
  Counting-process rows contribute `H(stop) - H(start)` for Weibull
  and multiphase.  The previous `hazard()` guard against non-zero
  starts is gone.
- **Delta-method prediction confidence limits** — shipped v0.9.8.
  Use `predict(..., se.fit = TRUE, level = 0.95)` to get a data frame
  with `fit`, `se.fit`, `lower`, and `upper` per row.  Closed-form
  Jacobian for Weibull and multiphase, `numDeriv::jacobian` fallback
  for exponential / log-logistic / log-normal.

---

## Rcpp acceleration note

`TemporalHazard` is currently pure R. If profiling against large real datasets
reveals bottlenecks in the likelihood kernel or optimizer inner loop, those
specific functions will be re-implemented with Rcpp. The public interface
(`hazard()`, `predict.hazard()`, etc.) will not change.


## References {.unnumbered}

Blackstone EH, Naftel DC, Turner ME Jr. The decomposition of time-varying
hazard into phases, each incorporating a separate stream of concomitant
information. *J Am Stat Assoc.* 1986;81(395):615–624.
doi: [10.1080/01621459.1986.10478314](https://doi.org/10.1080/01621459.1986.10478314)
